---
title: "Cleaning Our Data"
author: "Prof. Eric Friedlander"
footer: "[🔗 MATH 2025 - Fall 2026 -  Schedule](https://ericfriedlander.github.io/math2025-fa26/schedule)"
logo: "../images/logo.png"
format:
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk:
    R.options:
    width: 200
bibliography: references.bib
---


## Topics

-   Introduction to basic cleaning teachniques
    - Mutating our data
    - Filtering filtering our data

# Data

## DC Bikeshare

Our data set contains daily rentals from the Capital Bikeshare in Washington, DC in 2011 and 2012. It was obtained from the `dcbikeshare` data set in the `dsbox` R package.

We will focus on the following variables in the analysis:

-   `count`: total bike rentals
-   `temp_orig`: Temperature in degrees Celsius
-   `season`: 1 - winter, 2 - spring, 3 - summer, 4 - fall

Click [here](https://tidyverse.github.io/dsbox/reference/dcbikeshare.html) for the full list of variables and definitions.


## Loading Data From A File {.smaller}

- PATH: directions to get to a file
    - Example: "ae/ae-03/dataset.csv" will take it you to `ae` then `ae-03` then `dataset.csv`
- Two types of paths: absolute and relative
    - Absolute: starts from the root directory (i.e. the top-level directory)
    - Relative: starts from the current directory (`..` goes up one level)
- RStudio & Quarto: use `read_csv("relative_path/to/file.csv")` to load data from a CSV file
    - Brief aside: one of the few things we put in quotes in R. Use quotes when:
        - You want R to look outside of R for something
        - You are defining a string (i.e. a "word" in R)

## Reminder: Group Work Roles {.smaller}

Any time we are working on AEs, I will randomly assign you to groups of two or three. Each person will have a role:

- **Coder:**
    +   Shares screen and is the only person to type.
    +   Runs code, saves work, and makes the submission.
    +   Reads comments back when asked and asks for clarification when unsure.
- **Developer:**
    +   Directs the next steps: explains intent, outlines logic, and points out likely errors.
    +   Watches output, reads error messages aloud, and suggests targeted fixes.
- **Communicator:**
    +   Looks up help when needed.
    +   Reports group answers to the class.

## Application exercise

::: appex
📋 [AE 03 - DC Bikeshare](../ae/ae-03-bikeshare.qmd)
:::

Complete Exercises 1-3.

# Introduction to Simple Linear Regression

## Function between response and predictors

::: goal
$$\text{count} = f(\text{temp_orig}) + \epsilon$$
:::

-   **Goal**: Determine $f$
-   How do we determine $f$?
    -   Make an assumption about the functional form $f$ (parametric model)
    -   Use the data to fit a model based on that form

## Determine $f$

1.  Choose the functional form of $f$, i.e., **choose the appropriate model given the response variable**

-   Suppose $f$ takes the form of a linear model\
    $$y = f(\mathbf{X}) = \beta_0 + \beta_1 x_1 + \epsilon$$

. . .

2.  Use the data to fit (or train) the model, i.e, **estimate the model parameters,** $\beta_0, \beta_1$

## Computational set up {.smaller}

```{r}
#| echo: true
#| message: false

# load packages
library(tidyverse)       # for data wrangling
library(ggformula)       # for plotting
library(broom)           # for formatting model output
library(knitr)           # for formatting tables
```

```{r}
#| include: false
bikeshare <- read_csv(here::here("data/dcbikeshare.csv"))


# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%"
)
```


## Count vs. Temperature

```{r echo = F}
ggplot(data = bikeshare, aes(x = temp_orig, y = count)) +
  geom_point(alpha = 0.5) +
  labs(x = "Temperature (°C)",
      y  = "Count",
      title = "Bike Rentals vs. Temperature")
```

# Adding a third variable to our plots

## Skipping ahead to Exercise 5

- Suppose we want to add a third variable to our plots, such as the `season` variable. We have a few options:
    + Color the points based on the third variable (add `color = ~season` to the `gf_point` call)
    + Facet the plots by on the third variable by changing the formula to `count ~ temp_orig | season`

## Solution

```{r}
#| echo: TRUE

gf_point(count ~ temp_orig | season, data = bikeshare) |>
  gf_labs(x = "Temperature (Celsius)",
          y = "Daily bike rentals")
```

- What do we notice about how `season` is displayed


# Data cleaning

## `dplyr` and the `tidyverse`

- As R is an open-source programming language there are several difference styles and ways of doing things
- In this class we're going to use something called the `tidyverse`
- The `tidyverse` is a collection of R packages that share a common design philosophy
- `dplyr` is a package within the `tidyverse` that provides a consistent interface for data manipulation

## `dplyr` verbs {.smaller}

- `dplyr` provides a set of "verbs" for data manipulation
- We'll going to cover two today:
    - `mutate`: creates new columns and changes (mutates) existing columns
    - `filter`: selects rows based on a condition
- Your HW will go over two more:
    - `summarize`: compute summary statistics (e.g. mean, median, standard deviation)
    - `group_by`: group data by one or more variables (use this before `summarize` usually)
- We'll typically chain together these verbs using the pipe operator `|>` (you may see `%>%` in older code)
- Let's use these verbs to clean up the `bikeshare` data
- Reminder: [R Resources Tab in Sidebar](https://math2025-fa26.github.io/computing-r-resources/)

## Data prep {.smaller}

-   R calls categorical data "factors"
    - You may also see `<chr>` which stands for character. These are also categorical variables, but it's almost always better to convert them to factors
-   Factors remember what all possible levels (categories) and store them in order
    - You can put them in an order that makes sense so your plots looks better
-   Exercise 4: Recode `season` as a factor with names instead of numbers (livecode)

## Exercise 4 {.smaller}

-   Idea:
    +   Take the `bikeshare` data... AND THEN
    +   Use `case_when` to mutate the `season` column... AND THEN
    +   Use `factor` to convert `season` to a factor with levels in an order that makes sense

## Solution

```{r}
bikeshare <- bikeshare |>
  mutate(season = case_when(
    season == 1 ~ "winter",
    season == 2 ~ "spring",
    season == 3 ~ "summer",
    season == 4 ~ "fall"
  ),
  season = factor(season, levels = c("winter", "spring", "summer", "fall")))
```

## Back to Exercise 5 & 6

```{r}
#| echo: TRUE
#| fig-align: center

gf_point(count ~ temp_orig | season, data = bikeshare) |>
  gf_labs(x = "Temperature (Celsius)",
          y = "Daily bike rentals")
```

- Does it look like the relationship between `temp_orig` and `count` is the same in each season?
- Which is strongest/weakest?

## `season` is an important predictor

- Later in the course we'll learn how to include both `temp_orig` and `season` in our model
- For now, let's just focus on fitting a model to data from one season

# Filtering our data

## Filtering {.smaller}

::::{.columns}
:::{.column}
| ID | Thing1 | Thing2 | Thing3
--- | ------ | ------ | ------
1   | 100   | 200   | "Bleg"
2   | 20     | 20    | "Bleg"
3   | 30     | 40    | "Fleg"
4   | 120    | 120   | "Fleg"
:::
:::{.column}
- To keep some observations use `filter(`\[insert true/false statement here\]`)` and filter will keep all rows where the inside is true.
Examples:
    - `Thing2 <= 100`: keeps all rows where `Thing2` is less than or equal to 1000
    - `Thing3 == "Fleg"`: keeps everything where `Thing3` equals `"Fleg"`? (Always write strings in quotes)
    - `Thing1 != Thing2`: keeps everything where `Thing1` isn't equal to `Thing2`?
:::
::::

## Activity time

- Let's divide up the seasons
- I'm doing fall
- Do Exercises 7 & 8

## Sample Solution

```{r}
fall <- bikeshare |>
  filter(season == "fall")
```


## Rentals vs Temperature

**Goal**: Fit a line to describe the relationship between the temperature and the number of rentals in your season.

```{r}
#| echo: false
#| fig-align: center
p <- ggplot(data = fall, mapping = aes(x = temp_orig, y = count)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(x = "Temperature (Celsius)",
       y = "Daily bike rentals")

p
```

## Recap

::: {.incremental style="font-size: 0.85em"}
- Mutate: create and change columns
- Filter: keeps some of your observations
- `|>`: the `dplyr` pipe for chaining together data manipulation techniques
- Factors: R's way of remembering categorical data
:::
