---
title: "Simple Linear Regression"
author: "Prof. Eric Friedlander"
footer: "[🔗 MATH 2025 - Fall 2026 -  Schedule](https://ericfriedlander.github.io/math2025-fa26/schedule)"
logo: "../images/logo.png"
format:
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk:
    R.options:
    width: 200
bibliography: references.bib
---

## Application exercise

::: appex
📋 [AE 03 - Dr. F's Coffee](../ae/ae-03-coffee.qmd)
:::

Complete Exercises 1 and 2.

# Introduction to Simple Linear Regression

## Topics

-   Use simple linear regression to describe the relationship between a quantitative predictor and quantitative response variable.

-   Estimate the slope and intercept of the regression line using the least squares method.

## Computation set up {.smaller}

```{r}
#| echo: true
#| message: false

# load packages
library(tidyverse)       # for data wrangling
library(ggformula)       # for plotting
library(broom)           # for formatting model output
library(knitr)           # for formatting tables
library(janitor)         # for cleaning column names

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%"
)
```


# Data

## Dr. F's Coffee

Our data set is Dr. F's own coffee log: every cup of coffee he's brewed at home since January 2026, along with the beans, brewing method, and his own ratings of how the cup turned out. Each row is a single cup of coffee.

We will focus on the following variables in the analysis:

-   `vibes_rating`: Dr. F's overall rating of the cup, 0-10
-   `body`: sub-score for the cup's "body" (mouthfeel), 0-10
-   `roast`: the roast level of the beans (`Light`, `Light-Medium`, `Medium`, `Dark`)

::: {.callout-note}
There's also an `average` column, calculated from `body` and four other flavor sub-scores. Since it's just a summary of variables we already have, we won't use it here, same as in HW 01.
:::

Let's complete Exercises 3-6 together

## Data prep {.smaller}

-   Exercise 3: Order `roast` as a factor and fix a typo in `origin` (livecode)
-   Remember:
    +   Think of `|>` as "and then"
    +   `mutate` creates new columns and changes (mutates) existing columns
    +   R calls categorical data "factors"

```{r}
#| message: false
coffee <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTI_1K-jHHdbpi11aB5tX3Avr2eRXULfLg9BoveWhId4soH1BAnIJWuu0CiqO7YiP1TVoqsGi6kLbli/pub?gid=0&single=true&output=csv") |>
  clean_names() |>
  mutate(
    origin = if_else(origin == "Columbia", "Colombia", origin),
    roast  = factor(roast, levels = c("Light", "Light-Medium", "Medium", "Dark"))
  ) |>
  drop_na(vibes_rating, body, roast)
```

## Exploratory data analysis (Exercise 4)

```{r}
#| echo: TRUE
#| warning: false

gf_jitter(vibes_rating ~ body | roast, data = coffee, width = 0.15, height = 0.05, alpha = 0.5) |>
  gf_labs(x = "Body",
          y = "Vibes rating")
```

## One relationship, four roasts

-   (Exercise 5) Does the relationship between `body` and `vibes_rating` look about the same across roast levels, or different?

. . .

-   Unlike a data set where the relationship really does change by group, here it looks **similar in every panel** (roughly the same upward trend for Light, Light-Medium, and Medium)
-   `Dark` only has 2 cups. That's too few to say anything about a relationship, in that panel or between roasts.
-   So instead of filtering down to one group, like we might do when groups genuinely differ, we'll fit **one line** using every cup

```{r}
#| echo: false
m <- lm(vibes_rating ~ body, data = coffee)
```

## Vibes vs Body

**Goal**: Fit a line to describe the relationship between the body of a cup and Dr. F's overall rating of it.

```{r}
#| echo: false
#| fig-align: center
p <- ggplot(data = coffee, mapping = aes(x = body, y = vibes_rating)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(x = "Body",
       y = "Vibes rating")

p
```

## Why fit a line?

We fit a line to accomplish one or both of the following:

. . .

::: {style="color: #799100"}
**Prediction**
:::

What rating would we expect for a cup with a body score of 8?

. . .

::: {style="color : #799100"}
**Inference**
:::

Is body a useful predictor of the rating? By how much is the rating expected to change for each additional point of body?

## Population vs. Sample

**Population:** The set of items or events that you're interested in and hoping (able) to generalize the results of your analysis to.

**Sample:** The set of items that you have data for.

**Representative Sample:** A sample that looks like a small version of your population.

**Goal:** Build a model from your sample which generalizes to your population.

::: {.callout-note}
Here, the population is *every cup of coffee Dr. F could ever brew*, and the sample is the 127 cups he's actually logged so far. Is that sample representative? What might make it not be?
:::

## Terminology

::: columns
::: {.column width="30%"}
-   **Response, *Y***: variable describing the outcome of interest

-   **Predictor, *X***: variable we use to help understand the variability in the response
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: center
p
```
:::
:::

## Regression model

**Regression model:** a function that describes the relationship between a quantitative response, $Y$, and the predictor, $X$ (or many predictors).

```{=tex}
\begin{aligned} Y &= \color{black}{\textbf{Model}} + \text{Error} \\[8pt]
&= \color{black}{\mathbf{f(X)}} + \epsilon \\[8pt]
&= \color{black}{\boldsymbol{\mu_{Y|X}}} + \epsilon \end{aligned}
```

## Regression model

::: columns
::: {.column width="30%"}
```{=tex}
\begin{aligned} Y &= \color{purple}{\textbf{Model}} + \text{Error} \\[8pt]
&= \color{purple}{\mathbf{f(X)}} + \epsilon \\[8pt]
&= \color{purple}{\boldsymbol{\mu_{Y|X}}} + \epsilon \end{aligned}
```
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: center
ggplot(data = coffee, mapping = aes(x = body, y = vibes_rating)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "purple", se = FALSE) +
  labs(x = "X" ,
       y = "Y") +
  theme(axis.text = element_blank(),
  axis.ticks.x = element_blank(),
  axis.ticks.y = element_blank())
```
:::
:::

$\mu_{Y|X}$ is the mean value of $Y$ given a particular value of $X$.

## Regression model

::: columns
::: {.column width="30%"}
$$
\begin{aligned} Y &= \color{purple}{\textbf{Model}} + \color{blue}{\textbf{Error}} \\[5pt]
&= \color{purple}{\mathbf{f(X)}} + \color{blue}{\boldsymbol{\epsilon}} \\[5pt]
&= \color{purple}{\boldsymbol{\mu_{Y|X}}} + \color{blue}{\boldsymbol{\epsilon}} \\[5pt]
 \end{aligned}
$$
:::

::: {.column width="70%"}
```{r}
#| echo: false
#| fig-align: center
ggplot(data = coffee, mapping = aes(x = body, y = vibes_rating)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "purple", se = FALSE) +
  geom_segment(aes(x=body, xend=body, y=vibes_rating, yend=predict(m, coffee)), color = "blue", linetype = "dotted") +
  labs(x = "X" ,
       y = "Y") +
  theme_bw() +
  theme(axis.text = element_blank(),
  axis.ticks.x = element_blank(),
  axis.ticks.y = element_blank())
```
:::
:::

# Simple linear regression (SLR)

## SLR: Statistical model

- **Simple linear regression:** model to describe the relationship between $Y$ and $X$ where:
  + $Y$ is a quantitative/numerical response
  + $X$ is a *single* quantitative predictor
  + $$\Large{Y = \mathbf{\beta_0 + \beta_1 X} + \epsilon}$$

. . .

-   $\beta_1$: True slope of the relationship between $X$ and $Y$
-   $\beta_0$: True intercept of the relationship between $X$ and $Y$
-   $\epsilon$: Error

## SLR: Regression equation

$$\Large{\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X}$$

-   $\hat{\beta}_1$: Estimated slope of the relationship between $X$ and $Y$
-   $\hat{\beta}_0$: Estimated intercept of the relationship between $X$ and $Y$
-   $\hat{Y}$: Predicted value of $Y$ for a given $X$
-   No error term!

## Choosing values for $\hat{\beta}_1$ and $\hat{\beta}_0$

```{r}
#| echo: false
#| fig-align: center
ggplot(data = coffee, mapping = aes(x = body, y = vibes_rating)) +
  geom_point(alpha = 0.4) +
  geom_abline(intercept = 3.03, slope = 0.55, color = "red") +
  geom_abline(intercept = 5.5, slope = 0.15, color = "green") +
  geom_abline(intercept = 1.0, slope = 0.95, color = "blue") +
  labs(x = "Body",
       y = "Vibes rating")
```

## Residuals

```{r}
#| echo: false
#| fig-align: center
ggplot(data = coffee, mapping = aes(x = body, y = vibes_rating)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method ="lm", color = "red", se = FALSE) +
  geom_segment(aes(x=body, xend=body, y=vibes_rating, yend=predict(m, coffee)), color = "blue", linetype = "dotted") +
  labs(x = "Body",
       y = "Vibes rating") +
  theme(legend.position = "none")
```

$$\text{residual} = \text{observed} - \text{predicted} = y_i - \hat{y}_i$$

## Least squares line

-  **Residual** for the $i^{th}$ observation:

$$e_i = \text{observed} - \text{predicted}
= y_i - \hat{y}_i$$

-   **Sum of squared residuals**:

$$e^2_1 + e^2_2 + \dots + e^2_n$$

-   **Least squares line** is the one that minimizes the sum of squared residuals

```{r}
#| echo: false
sx <- round(sqrt(var(coffee$body)),4)
sy <- round(sqrt(var(coffee$vibes_rating)),4)
r <- round(cor(coffee$body, coffee$vibes_rating),4)
xbar <- round(mean(coffee$body),4)
ybar <- round(mean(coffee$vibes_rating),4)
```

::: appex
Finish Exercises 7-11 on your own.
:::

## Recap

::: {.incremental style="font-size: 0.85em"}

- Simple linear regression (SLR) describes the relationship between a quantitative predictor ($X$) and response ($Y$).
- The regression equation: $\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X$
- $\hat{\beta}_1$ (slope): estimated change in $Y$ for each unit increase in $X$
- $\hat{\beta}_0$ (intercept): estimated value of $Y$ when $X = 0$
- Residuals: difference between observed and predicted values ($e_i = y_i - \hat{y}_i$)
- Least squares method chooses coefficients to minimize the sum of squared residuals
- SLR can be used for prediction and inference
:::

## For Monday

- Complete and submit AE 03
- Completed readings

## References
