---
title: "Simple Linear Regression"
author: "Prof. Eric Friedlander"
footer: "[🔗 MATH 2025 - Fall 2026 -  Schedule](https://ericfriedlander.github.io/math2025-fa26/schedule)"
logo: "../images/logo.png"
format:
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk:
    R.options:
    width: 200
bibliography: references.bib
---

```{r}
#| include: false
#| message: false

# load packages
library(tidyverse)       # for data wrangling
library(ggformula)       # for plotting
library(broom)           # for formatting model output
library(knitr)           # for formatting tables
bikeshare <- read_csv(here::here("data/dcbikeshare.csv"))


# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%"
)

bikeshare <- read_csv("../data/dcbikeshare.csv") |>
  mutate(season = case_when(
    season == 1 ~ "winter",
    season == 2 ~ "spring",
    season == 3 ~ "summer",
    season == 4 ~ "fall"
  ),
  season = factor(season))
```


## Topics

-   Use simple linear regression to describe the relationship between a quantitative predictor and quantitative response variable.
-   Estimate the slope and intercept of the regression line using R.

# Data

## DC Bikeshare

Our data set contains daily rentals from the Capital Bikeshare in Washington, DC in 2011 and 2012. It was obtained from the `dcbikeshare` data set in the `dsbox` R package.

We will focus on the following variables in the analysis:

-   `count`: total bike rentals
-   `temp_orig`: Temperature in degrees Celsius
-   `season`: 1 - winter, 2 - spring, 3 - summer, 4 - fall

Click [here](https://tidyverse.github.io/dsbox/reference/dcbikeshare.html) for the full list of variables and definitions.

## Last time {.smaller}

::::{.columns}
:::{.column}
- Coloring and faceting our plots
- Used `mutate` to recode `season` as a `factor` with `winter`, `fall`, `summer`, `spring`
- Talked about how to use `filter` to select certain observations
- Noticed that the relationship between `count` and `temp_orig` didn't really look linear:
    - About to use `filter` for each group to select a season
:::
:::{.column}
```{r}
#| echo: false
ggplot(data = bikeshare, aes(x = temp_orig, y = count)) +
  geom_point(alpha = 0.5) +
  labs(x = "Temperature (°C)",
      y  = "Count",
      title = "Bike Rentals vs. Temperature")
```
:::
::::

# Filtering our data

## Filtering {.smaller}

::::{.columns}
:::{.column}
| ID | Thing1 | Thing2 | Thing3
--- | ------ | ------ | ------
1   | 100   | 200   | "Bleg"
2   | 20     | 20    | "Bleg"
3   | 30     | 40    | "Fleg"
4   | 120    | 120   | "Fleg"
:::
:::{.column}
- To keep some observations use `filter(`\[insert true/false statement here\]`)` and filter will keep all rows where the inside is true.
Examples:
    - `Thing2 <= 100`: keeps all rows where `Thing2` is less than or equal to 1000
    - `Thing3 == "Fleg"`: keeps everything where `Thing3` equals `"Fleg"`? (Always write strings in quotes)
    - `Thing1 != Thing2`: keeps everything where `Thing1` isn't equal to `Thing2`?
:::
::::

## Activity time: Picking up where we left off

::: appex
📋 [AE 03 - DC Bikeshare](../ae/ae-03-bikeshare.qmd)
:::

- Get in same groups as last time
- Rotate rolls if you feel comfortable with someone else typing on your computer (Coder -> Communicator -> Developer -> Coder)
- Let's divide up the seasons
- I'm doing fall
- Do Exercises 7 & 8
    - Note: Exercise 7 is a challenge that is unrelated to choosing a season

## Sample Solution For 8

```{r}
fall <- bikeshare |>
  filter(season == "fall")
```


## Rentals vs Temperature

**Goal**: Fit a line to describe the relationship between the temperature and the number of rentals in your season.

```{r}
#| echo: false
#| fig-align: center
p <- ggplot(data = fall, mapping = aes(x = temp_orig, y = count)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(x = "Temperature (Celsius)",
       y = "Daily bike rentals")

p
```

# Simple linear regression (SLR)

## SLR: Statistical model

- **Simple linear regression:** model to describe the relationship between $Y$ and $X$ where:
  + $Y$ is a quantitative/numerical response
  + $X$ is a *single* quantitative predictor
  + $$\Large{Y = \mathbf{\beta_0 + \beta_1 X} + \epsilon}$$

. . .

-   $\beta_1$: True slope of the relationship between $X$ and $Y$
-   $\beta_0$: True intercept of the relationship between $X$ and $Y$
-   $\epsilon$: Error

## SLR: Regression equation

$$\Large{\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X}$$

-   $\hat{\beta}_1$: Estimated slope of the relationship between $X$ and $Y$
-   $\hat{\beta}_0$: Estimated intercept of the relationship between $X$ and $Y$
-   $\hat{Y}$: Predicted value of $Y$ for a given $X$
-   No error term!

## Statistical model vs. regression equation {.midi}

**Statistical model (also known as data-generating model)**

$${\small \text{count} = \beta_0 + \beta_1 ~\text{temp_orig} + \epsilon}$$

Models the process for generating values of the response in the population (function + error)

<br>

. . .

**Regression equation**

Estimate of the function using the sample data

$${\small \hat{\text{count}} = \hat{\beta}_0 + \hat{\beta}_1 ~\text{temp_orig}}$$

## Why fit a model?

-   **Prediction:** Expected value of the response variable for given values of the predictor variables

-   **Inference:** Conclusion about the relationship between the response and predictor variables

::: question
-   What is an example of a **prediction** question that can be answered using the model of count vs. temperature?

-   What is an example of an **inference** question that can be answered using the model of count vs. temperature?
:::


## Choosing values for $\hat{\beta}_1$ and $\hat{\beta}_0$

```{r}
#| echo: false
#| fig-align: center
ggplot(data = fall, mapping = aes(x = temp_orig, y = count)) +
  geom_point(alpha = 0.4) +
  geom_abline(intercept = -111.0, slope = 222.4, color = "red") +
  geom_abline(intercept = 1000, slope = 100, color = "green") +
  geom_abline(intercept = -1000, slope = 300, color = "blue") +
  labs(x = "Temperature (Celsius)",
       y = "Daily bike rentals")
```

# Estimating the regression line in R

## Fit model & estimate parameters

```{r}
#| echo: true
#| code-line-numbers: "|1"

fall_fit <- lm(count ~ temp_orig, data = fall)
fall_fit
```

## Look at the regression output

```{r}
#| echo: true
#| code-line-numbers: "|2"

fall_fit <- lm(count ~ temp_orig, data = fall)
fall_fit
```


$$\widehat{\text{count}} = 2077.5 + 152.9 \times \text{temp_orig}$$

## Use `tidy` to make it prettier

We'll focus on the first column for now...

```{r}
#| echo: true
#| code-line-numbers: "|2"

fall_fit |>
  tidy()
```

## Using `kable` to make it extra pretty


```{r}
#| echo: true
#| code-line-numbers: "|3"

fall_fit |>
  tidy() |>
  kable()
```

## Visualize Model

```{r}
#| echo: true
#| code-fold: true

fall |>
  gf_point(count ~ temp_orig) |>
  gf_lm() |>
  gf_labs(x = "Temperature (°C)", y = "Daily Bike Rentals", title = "Fall Bike Rentals vs. Temperature")
```

# Interpretation

## Interpreting the Slope and Intercept

-  Slope: For each additional unit of $X$ we expect $Y$ to increase by $\hat{\beta}_1$, on average.
-  Intercept: If $X$ were 0, we predict $Y$ to be $\hat{\beta}_0$

## Fall Example

- Slope: For each addition degree Celcius we expect the number of bike rentals to increase by `r round(fall_fit$coefficients[2], 2)` on average, in the Fall.
- Intercept: If the temperature were 0 degrees Celcius, we predict the number of bike rentals to be `r round(fall_fit$coefficients[1], 2)` in the Fall.


## Does it make sense to interpret the intercept?

`r emo::ji("white_check_mark")` **The intercept is meaningful in the context of the data if**

-   the predictor can feasibly take values equal to or near zero, or

-   there are values near zero in the observed data.

. . .

`r emo::ji("stop_sign")` Otherwise, the intercept may not be meaningful!

-   Note meaningful doesn't mean wrong

. . .

Complete the remainder of the exercises.

# Prediction

## Fall Model

```{=tex}
\begin{aligned}
\widehat{Y} &=  2077.48 + 152.87 \times X\\

\widehat{\text{count}} &=  2077.48 + 152.87 \times \text{temp_orig}
\end{aligned}
```

. . .

On the white board: Suppose that it's 15 degrees Celsius outside. According to this model, how many bike rentals should we expect if it's your season?

## Making a prediction

On the white board: Suppose that it's 15 degrees Celsius outside. According to this model, how many bike rentals should we expect if it's your season?

```{=tex}
\begin{aligned}

\widehat{\text{count}} &=  2077.48 + 152.87 \times \text{temp_orig}\\
&= 2077.48 + 152.87 \times 15 \\
&= 4370.53
\end{aligned}
```

About 4,371 rentals!

## Prediction in R

- `predict` function takes two arguments
    - First: a model that's been fit
    - Second: a new dataset with the **same column names** as the dataset used to train the model

## Examples

::::{.columns}
:::{.column}
```{r}
#| echo: true

fall_predictions <- predict(fall_fit, fall)
fall_predictions |>
  head() |> # what does this do?
  kable()
```
:::
:::{.column}
```{r}
#| echo: true

# create a data frame for a new temperature
new_days <- tibble(temp_orig = c(15, 20, 5, 10))

# I don't need to save the predictions
predict(fall_fit, new_days) |>
  kable()
```
:::
::::

## Exercise 14

1. Add a new exercise to your activity (you can start by copy pasting a previous exercise and answer block).
2. Confirm you answer to the white board question using the `predict` in R. I.e. what would you model predict the number of bike rentals to be if it was 15 degrees out.

## Recap

::: {.incremental style="font-size: 0.85em"}
- Simple linear regression (SLR) describes the relationship between a quantitative predictor ($X$) and response ($Y$).
- The statistical model: $Y = \beta_0 + \beta_1 X$
- The regression equation: $\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X$
- $\hat{\beta}_1$ (slope): estimated change in $Y$ for each unit increase in $X$
- $\hat{\beta}_0$ (intercept): estimated value of $Y$ when $X = 0$
- SLR can be used for prediction and inference
:::
