---
title: "Categorical Predictors"
author: "Prof. Eric Friedlander"
footer: "[🔗 MATH 2025 - Fall 2026 - Schedule](https://ericfriedlander.github.io/math2025-fa26/schedule)"
logo: "../images/logo.png"
format:
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false
    chalkboard: true
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk:
    R.options:
    width: 200
bibliography: references.bib
---

```{r}
#| include: false
#| message: false

# load packages
library(tidyverse)       # for data wrangling
library(ggformula)       # for plotting
library(mosaic)          # for favstats
library(broom)           # for formatting model output
library(knitr)           # for formatting tables
library(coursekata)      # for gf_model

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

# set default figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%"
)
```

## Topics

- Use regression to describe the relationship between a **categorical** predictor and a quantitative response variable.
- Understand indicator (dummy) variables and reference levels.
- Interpret the intercept and slope when the predictor is categorical.

# Last time

## Last time {.smaller}

::::{.columns}
:::{.column}

- Fit `lm(count ~ temp_orig)` for our bikeshare data
- Interpreted the slope and intercept
- Used `predict()` to make predictions
- Every predictor so far has been a **number**
  :::
  :::{.column}

```{r}
#| echo: false
bikeshare <- read_csv("../data/dcbikeshare.csv") |>
  mutate(season = case_when(
    season == 1 ~ "winter",
    season == 2 ~ "spring",
    season == 3 ~ "summer",
    season == 4 ~ "fall"
  ),
  season = factor(season, levels = c("winter", "spring", "summer", "fall")))

fall <- bikeshare |> filter(season == "fall")
fall_fit <- lm(count ~ temp_orig, data = fall)

fall |>
  gf_point(count ~ temp_orig) |>
  gf_lm() |>
  gf_labs(x = "Temperature (°C)", y = "Daily Bike Rentals")
```

:::
::::

## What if the predictor isn't a number? {.smaller}

- We also have `season`: winter, spring, summer, fall.
- What would `lm(count ~ season)` even mean?

::: question

- You can't multiply "fall" by a slope. So what does $\hat{\beta}_1$ look like when $X$ is a category?
- Can we still write $\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X$?

:::

# Meet the Coffee Truck

## The scenario {.smaller}

Jo inherited a coffee truck from an uncle. After repairs, Jo has **\$20,000** left. Coffee costs **\$0.50/cup** to make, and fixed daily expenses are **\$300/day**. Jo wants help figuring out what combination of settings sells the most coffee.

We'll use the [Coffee Truck game](https://www.stat2games.sites.grinnell.edu/games/coffeetruck.html) (Grinnell **stat2games**) to _generate our own data_ by playing simulated shifts.

## Settings you can choose {.smaller}

Before each shift you pick a combination of:

| Setting  | Options in the game                                   |
| -------- | ----------------------------------------------------- |
| Location | Business District, City Park, City Hall, Near the Zoo |
| Music    | No Music, Hip Hop, Alternative                        |
| Price    | \$2, \$3, \$4, \$5 per cup                            |
| Time     | Morning, Lunch, Afternoon, Evening                    |
| Weather  | Relatively Constant, Variable                         |

**Response variable:** `Sales` -- the number of cups sold that shift.

## Our baseline {.smaller}

Everyone in class will start from the same baseline:

> **Business District**, **No Music**, **\$3 per cup**, **Lunch**

Each group then picks **one setting** to change to a second level, holding everything else at the baseline. (If there's time, we'll add a third level of that same setting later.)

::: appex
📋 [AE 05 - The Coffee Truck](../ae/ae-05-coffee-truck.qmd)
:::

- I'm doing **Location** (Business District vs. City Park); your group picks one other setting to change
- Complete Exercises 1-7

# Regression with one categorical predictor

## We're just encoding means

What comes next may seem complicated, but we're just building a model that predicts the groups means:

```{r}
bikeshare |>
  group_by(season) |>
  summarize(mean_count = mean(count)) |>
  kable(digits = 3)
```

## Indicator variables {.smaller}

- Suppose a categorical variable has $K$ categories (levels).
- We can make $K$ **indicator variables**, one per category.
- An indicator variable takes the value 1 or 0:
  - 1 if the observation belongs to that category
  - 0 if it does not
- Finally... we drop one of the indicator variables.
- This category is called the **reference level** (or **baseline**) and it doesn't get its own term in the model

## Example {.smaller}

::::{.columns}
:::{.column}
Create indicator variables for `Season` in our bikeshare data:

```{r}
#| echo: false
set.seed(2024)
bike_slice <- bikeshare |>
  mutate(
    winter = if_else(season == "winter", 1, 0),
    spring = if_else(season == "spring", 1, 0),
    summer = if_else(season == "summer", 1, 0),
    fall = if_else(season == "fall", 1, 0)
  ) |>
  select(season, winter, spring, summer, fall) |>
  slice_sample(n=8)
 bike_slice |>
  kable()
```

:::
:::{.column}
Set `winter` as the reference level, and drop it from the model:

```{r}
#| echo: false
bike_slice |>
  select(season, spring, summer, fall) |>
  kable()
```

:::
::::

## Indicators in the model {.smaller}

- We only use $K - 1$ of the indicator variables in the model.
- The **reference level** (or **baseline**) is the category left out — it doesn't get its own term.
- The coefficient on each indicator is the expected _difference_ in the response compared to the baseline.
- This is called **dummy coding**, and _R does it for you_.
- But why... (Board work)

## R's default reference level {.smaller}

By default, R picks the reference level either alphabetically (if variable is not a factor) or the first level (if variable is a factor).

Most of the time, you probably don't care which level is chosen to be the reference level (you need to know which one it is though)... but if you do chage it using the same trick from `season` in Lecture 3:

```{r}
#| eval: false
bikeshare <- bikeshare |>
  mutate(season = factor(season, levels = c("winter", "spring", "summer", "fall")))
```

## Fitting the model {.smaller}

::::{.columns}
:::{.column}

```{r}
season_mod <- lm(count ~ season, data = bikeshare)
season_mod |>
  tidy() |>
  kable(digits = 3)
```

:::
:::{.column}

```{r}
#| echo: false
bikeshare |>
  gf_jitter(count~ season, height = 0, width = 0.15) |>
  gf_model(season_mod, color="red") |>
  gf_labs(y = "Daily Bike Rentals")
```

:::
::::

$$
\widehat{\text{count}} = \hat\beta_0 + \hat\beta_1 \times \text{seasonspring} + \hat\beta_2 \times \text{seasonsummer} + \hat\beta_3 \times \text{seasonfall}
$$

::: question

- Where can we figure out what the reference level is from the output?
- Where do you see the intercept and slope in the plot?
- What does this model predict for a winter day? A summer day?

:::

## The intercept is a mean; the slope is a difference {.smaller}

::::{.columns}
:::{.column}

```{r}
season_mod |> tidy() |> kable(digits = 3)
```

:::
:::{.column}

```{r}
mean(count ~ season, data = bikeshare) |>
  kable(digits = 3)
```

:::
::::

- **Intercept**: we predict a winter (baseline) day to have $\hat\beta_0$ rentals, on average — this _is_ the winter mean.
- **Slope**: switching to summer, we expect $\hat\beta_1$ more rentals, on average, compared to winter — this _is_ the difference between the two means.

. . .

"A one-unit increase in $X$" now means "switching from the baseline category to this one."

## (Trick question) {.smaller}

Are `season` and `count` **correlated**?

. . .

- Correlation only makes sense for two _quantitative_ variables.
- `season` is categorical — there's no correlation coefficient to compute, even though we can absolutely fit a regression model.

::: appex
Complete Exercises 6-13.
:::

::: appex
If time permits, complete Exercises 14-15.
:::

# Wrap up

## Recap

::: {.incremental style="font-size: 0.85em"}

- Regression works with a categorical predictor $X$, just like a quantitative one.
- R recodes a $K$-level categorical variable into $K - 1$ **indicator (dummy) variables**.
- The left-out category is the **reference level**; by default R picks it alphabetically, but we can set it with `factor(x, levels = ...)`.
- The **intercept** is the mean of the response at the reference level.
- Each **slope** is the difference in means between that level and the reference level.
- Correlation doesn't apply to categorical predictors, but regression still does.
  :::
