---
title: "Residuals, Least Squares, and Model Evaluation"
author: "Prof. Eric Friedlander"
footer: "[🔗 MATH 2025 - Fall 2026 - Schedule](https://ericfriedlander.github.io/math2025-fa26/schedule)"
logo: "../images/logo.png"
format:
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk:
    R.options:
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)
```

## Topics

- Define residuals, and connect them to the idea of variation in $Y$.
- Understand how the least squares method chooses a regression line.
- Partition variation into model and error (ANOVA), and use it to compute $R^2$ and RMSE.

## Last time {.smaller}

::: {.incremental style="font-size: 0.9em"}
- Regression works with a categorical predictor $X$, just like a quantitative one.
- R recodes a $K$-level categorical variable into $K - 1$ **indicator (dummy) variables**, dropping the **reference level**.
- The intercept is the mean of the response at the reference level; each slope is the difference in means from that level.
- We started playing with the **Coffee Truck** to generate our own data.
:::

## Computational set up

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(ggformula)   # for plotting using formulas
library(broom)       # for formatting model output
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(kableExtra)  # also for pretty tables
library(patchwork)   # arrange plots
library(yardstick)   # for R^2 and RMSE

# Spotify Dataset
spotify <- read_csv("../data/spotify-popular.csv")

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

## Quick Data Cleaning

```{r}
#| echo: true

spotify <- spotify |>
  mutate(duration_min = duration_ms / 60000)
```

::: question
- What is this code doing?
- Why might I be doing it?
:::

## The regression model, revisited {.smaller}

```{r}
#| echo: true
spotify_fit <- lm(danceability ~ duration_min, data = spotify)

tidy(spotify_fit, conf.int = TRUE, conf.level = 0.95) |>
   kable(digits = 3)
```

- There is strong statistical evidence that there is a linear relationship between the duration of a song and its danceability.
- We are 95% confident that as the length of a song increases by one minute the danceability will decrease by between 0.009 and 0.039 units.

# Residuals

## From points to a line {.smaller}

```{r}
#| echo: false
set.seed(2025)
spotify_sample <- spotify |>
  slice_sample(n = 30)

ggplot(data = spotify_sample, mapping = aes(x = duration_min, y = danceability)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", color = "red", se = FALSE, fullrange = TRUE) +
  labs(x = "Duration (minutes)", y = "Danceability")
```

::: question
For any one song, how far off is the line from the actual data point?
:::

## Residuals {.smaller}

```{r}
#| echo: false
spotify_sample_aug <- augment(lm(danceability ~ duration_min, data = spotify_sample), data = spotify_sample)

ggplot(data = spotify_sample_aug, mapping = aes(x = duration_min, y = danceability)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", color = "red", se = FALSE, fullrange = TRUE) +
  geom_segment(aes(x = duration_min, xend = duration_min, y = danceability, yend = .fitted),
               color = "blue", linetype = "dotted") +
  labs(x = "Duration (minutes)", y = "Danceability")
```

$$\text{residual} = \text{observed} - \text{predicted} = y_i - \hat{y}_i = e_i$$

::: {.fragment}
- A **positive** residual: the point is *above* the line (model underpredicted).
- A **negative** residual: the point is *below* the line (model overpredicted).
:::

# Partitioning Variability

## Let's think about variation

:::{.incremental}
- DATA = MODEL + ERROR
- $\substack{\text{Variation} \\ \text{in Y}} = \substack{\text{Variation explained} \\ \text{by model}} + \substack{\text{Variation not explained} \\ \text{by model}}$
- Every point's residual is a leftover piece of variation the model didn't explain.
:::

::: question
If we drew *any* other line through the cloud of points, would the residuals get bigger or smaller? Is there a "best" line?
:::

# Least squares

## Choosing values for $\hat{\beta}_1$ and $\hat{\beta}_0$

```{r}
#| echo: false
#| fig-align: center
ggplot(data = spotify_sample, mapping = aes(x = duration_min, y = danceability)) +
  geom_point(size = 2) +
  geom_abline(intercept = 0.781, slope = -0.024, color = "red") +
  geom_abline(intercept = 0.9, slope = -0.06, color = "green") +
  geom_abline(intercept = 0.65, slope = 0.01, color = "blue") +
  labs(x = "Duration (minutes)", y = "Danceability")
```

::: question
Which of these three lines does the "best" job? How would you decide?
:::

## The least squares method {.smaller}

::: incremental
- We need a rule for picking *the* line, not just *a* line.
- **Least squares method**: choose $\hat{\beta}_0$ and $\hat{\beta}_1$ to minimize the sum of squared residuals
$$
\sum_{i=1}^n e_i^2 = \sum_{i=1}^n (y_i - \hat{y}_i)^2
$$
- Why square the residuals instead of just adding them up?
    + Positive and negative residuals would cancel out.
    + Squaring also penalizes large misses more than small ones.
:::

## Properties of least squares regression {.smaller}

::: incremental
-   Passes through the center of mass point, the coordinates corresponding to average $X$ and average $Y$: $\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1\bar{X}$

-   Slope has same sign as the correlation coefficient: $\hat{\beta}_1 = r \frac{s_Y}{s_X}$

    + $r$: correlation coefficient
    + $s_Y, s_X$: sample standard deviations of $X$ and $Y$

-   Sum of the residuals is zero: $\sum_{i = 1}^n e_i \approx 0$

    +   Intuition: Residuals are "balanced"

-   The residuals and $X$ values are uncorrelated
:::

## Estimating the slope, by hand {.smaller}

$$\large{\hat{\beta}_1 = r \frac{s_Y}{s_X}}$$

```{r}
#| echo: true
s_x <- sd(spotify$duration_min)
s_y <- sd(spotify$danceability)
r <- cor(spotify$duration_min, spotify$danceability)

s_x
s_y
r
```

::: columns
::: {.column width="50%"}
```{r}
#| echo: true
b1 <- r * s_y / s_x
b1
```
:::
::: {.column width="50%"}
Compare to `lm()`:
```{r}
#| echo: true
coef(spotify_fit)["duration_min"]
```
:::
:::

## Estimating the intercept, by hand {.smaller}

$$\large{\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1\bar{X}}$$

```{r}
#| echo: true
x_bar <- mean(spotify$duration_min)
y_bar <- mean(spotify$danceability)

b0 <- y_bar - b1 * x_bar
b0
```

Compare to `lm()`:

```{r}
#| echo: true
coef(spotify_fit)["(Intercept)"]
```

<br>

::: small
[Click here](https://introregression.netlify.app/98-appendix) for details on deriving the equations for slope and intercept, which is easy if you know multivariate calculus.
:::

# Partitioning Variability (ANOVA)

## Splitting up the residual {.smaller}

:::{.incremental}
-   $y_i - \bar{y} = (\hat{y}_i - \bar{y}) + (y_i-\hat{y}_i)$
-   Square and sum: $\sum(y_i-\bar{y})^2 = \sum(\hat{y} - \bar{y})^2 + \sum(y-\hat{y})^2$
-   $\substack{\text{Sum of squares} \\ \text{Total}} = \substack{\text{Sum of squares} \\ \text{model}} + \substack{\text{Sum of squares} \\ \text{error}}$
-   $SSTotal = SSModel + SSE$
-   $SST = SSM + SSE$
-   ANOVA: Analysis of Variance
:::

## ANOVA in R

```{r}
spotify_fit |>
  anova() |>
  tidy() |>
  kable()
```

## Recall: Correlation Coefficient

-   The **correlation coefficient**, $r$, is a number between -1 and +1 that measures how strong the linear relationship between two variables $x$ and $y$ is.

$$
r = \frac{\sum(x_i - \bar{x})(y_i-\bar{y})}{\sqrt{\sum(x_i-\bar{x})^2\sum(y_i-\bar{y})^2}}
= \frac{\sum(x_i - \bar{x})(y_i-\bar{y})}{s_xs_y}
$$

## Two statistics: $R^2$ {.smaller}

::: incremental
-   **R-squared**, $R^2$, **Coefficient of Determination**: Percentage of variability in the outcome explained by the regression model (in the context of SLR, the predictor)
$$
R^2 = Cor(y, \hat{y})^2
$$
    -  Also called **PRE (Percent Reduction in Error)** because:
$$
R^2 = \frac{SSModel}{SSTotal}
$$
:::

## $R^2$

::: incremental
-   Ranges between 0 (terrible predictor) and 1 (perfect predictor)
-   Has no units
-   Calculate with `rsq()` from `yardstick` package using the augmented data:
```{r}
#| echo: true

spotify_aug <- augment(spotify_fit)

rsq(spotify_aug, truth = danceability, estimate = .fitted) |> kable()
```
:::

## Interpreting $R^2$ {.smaller}

```{r}
#| echo: false
spotify_fit_rsq <- round(rsq(spotify_aug, truth = danceability, estimate = .fitted) |> pull(.estimate) * 100, 1)
```

::: poll
🗳️ **Discussion**

::: midi
The $R^2$ of the model for `danceability` from `duration_min` is `r spotify_fit_rsq`%. Which of the following is the correct interpretation of this value?
:::

1.  `duration_min` correctly predicts `r spotify_fit_rsq`% of `danceability`.
2.  `r spotify_fit_rsq`% of the variability in `danceability` can be explained by `duration_min`.
3.  `r spotify_fit_rsq`% of the variability in `duration_min` can be explained by `danceability`.
4.  `r spotify_fit_rsq`% of the time `danceability` can be predicted by `duration_min`.
:::

## Activity

::: appex
In groups, at the board, design a simulation-based procedure for producing a p-value for the following hypothesis test.

- $H_0: R^2 = 0$
- $H_A: R^2 \neq 0$
:::

## RMSE {.smaller}

::: incremental
-   **Root mean square error, RMSE**: A measure of the average error (average difference between observed and predicted values of the outcome)
$$
RMSE = \sqrt{\frac{\sum_{i = 1}^n (y_i - \hat{y}_i)^2}{n}}
$$
    + Sometimes people just care about numerator (SSE) or version without the square-root (MSE)
    + Sometimes the denominator may have $n-1$ instead
-   Ranges between 0 (perfect predictor) and infinity (terrible predictor)
-   Same units as the response variable
-   Interpretation (kind of): how much does my model miss by, on average.
:::

## RMSE in R

```{r}
#| echo: true

rmse(spotify_aug, truth = danceability, estimate = .fitted)
```

# Back to the Coffee Truck

## Comparing models with $R^2$ {.smaller}

Recall Jo's coffee truck: **Business District**, **Lunch**, **Weather = Constant** the whole time.

Today, each group will generate **three** data sets with the [Coffee Truck game](https://www.stat2games.sites.grinnell.edu/games/coffeetruck.html):

| Data set | What varies | What's held fixed |
|:--:|:--|:--|
| `music` | Music (No Music / Hip Hop / Alternative) | Price = \$3 |
| `price` | Price (\$2 / \$3 / \$4 / \$5) | Music = No Music |
| `both`  | Music **and** Price | — |

::: appex
📋 [AE 06 - Comparing Models with the Coffee Truck](../ae/ae-06-coffee-truck-eval.qmd)
:::

Complete Exercises 1-3.

```{r}
#| echo: false
countdown(minutes = 15)
```

## A data cleaning wrinkle {.smaller}

The game exports `Price` as **text**, e.g. `"$3"`, not a number:

```{r}
#| echo: false
tibble(Price = c("$2", "$3", "$4", "$5")) |> kable()
```

::: question
Why does this matter for fitting `lm(Sales ~ Price)`? What do we need to do first?
:::

. . .

Same fix as before: strip the `$` and convert.

```{r}
#| eval: false
#| echo: true
price <- price |>
  mutate(Price = str_replace(Price, "\\$", "")) |>
  mutate(Price = as.numeric(Price))
```

Complete Exercises 4-9.

## The comparison {.smaller}

You'll fit `Sales ~ Music` on **`music`**, and both `Sales ~ Music` and `Sales ~ Price` on **`both`**.

::: question
Music's effect on `Sales` didn't change between `music` and `both`. So why might $R^2$ for `Sales ~ Music` be so different in these two data sets?
:::

. . .

- In `music`, price is fixed — all the leftover variability in `Sales` is either explained by music or is pure noise.
- In `both`, price is also bouncing around. That extra variability inflates **SST**, so the *same* music effect now explains a *smaller share* of a *bigger* pie.

Complete Exercises 10-12. If time permits, complete Exercise 13.

## Using the word "Good"

-   There is no such thing as a "Good" $R^2$ or, especially, RMSE without context.
-   Whether your model is a "Good" model depends on many things:
    +   What are you using your model for?
    +   How good are other models?
    +   **What else was varying in the data you fit it to?**

## Recap {.smaller}

-   A **residual** is the leftover gap between an observed value and the model's prediction: $e_i = y_i - \hat{y}_i$.
-   The **least squares method** picks $\hat{\beta}_0, \hat{\beta}_1$ to minimize $\sum e_i^2$.
-   Can decompose total variation (SST) into variation explained by the model (SSM) and leftover variation (SSE).
-   Two metrics for evaluating and comparing models:
    +   $R^2$: What proportion of the variation in the response variable is explained by the model?
    +   $RMSE$: How far does my model miss by, on average?
-   $R^2$ depends on the data it was fit to, not just the true relationship — so it can't be compared across data sets that vary different things.
