---
title: "Residuals, Least Squares, and Model Evaluation"
author: "Prof. Eric Friedlander"
footer: "[🔗 MATH 2025 - Fall 2026 - Schedule](https://ericfriedlander.github.io/math2025-fa26/schedule)"
logo: "../images/logo.png"
format:
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk:
    R.options:
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

# load packages
library(tidyverse)   # for data wrangling and visualization
library(ggformula)   # for plotting using formulas
library(broom)       # for formatting model output
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(kableExtra)  # also for pretty tables
library(patchwork)   # arrange plots
library(yardstick)   # for R^2 and RMSE
library(supernova)   # for ANOVA
library(coursekata)

# DC Bikeshare Dataset
bikeshare <- read_csv(here::here("data/dcbikeshare.csv")) |>
  mutate(season = case_when(
    season == 1 ~ "winter",
    season == 2 ~ "spring",
    season == 3 ~ "summer",
    season == 4 ~ "fall"),
    workingday = if_else(workingday == 1, "Workingday", "Holiday"),
  season = factor(season, levels = c("winter", "spring", "summer", "fall")),
  workingday = factor(workingday))

fall <- bikeshare |> 
  filter(season == "fall")
# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

## Topics

- Define residuals, and connect them to the idea of variation in $Y$.
- Understand how the least squares method chooses a regression line.
- Partition variation into model and error (ANOVA), and use it to compute $R^2$ and RMSE.

## Last time {.smaller}

-   A **residual** is the leftover gap between an observed value and the model's prediction: $e_i = y_i - \hat{y}_i$.
-   The **least squares method** picks $\hat{\beta}_0, \hat{\beta}_1$ to minimize $\sum e_i^2$.
-   Can decompose total variation (SST) into variation explained by the model (SSM) and leftover variation (SSE).

## Empty Model (Mean) vs. Temperature Model

Consider the following two models:

::::{.columns}
:::{.column}
This model just predicts the mean and is called the **empty model**
```{r}
mean_model <- lm(count ~ NULL, data = fall)
mean_model |> tidy() |> kable()
```
:::
:::{.column}
This model uses `temp_orig` as a predictor and we'll call it the **temperature model**
```{r}
temp_model <- lm(count ~ temp_orig, data = fall)
temp_model |>  tidy() |> kable()
```
:::
::::

- Which model do you expect to have the smaller SSE?
- Which model will explain more variation in `count`?

## Let's think about variation {.smaller}

:::{.incremental}
- DATA = MODEL + ERROR
- $\substack{\text{Total Variation} \\ \text{in Y}} = \substack{\text{Variation explained} \\ \text{by model}} + \substack{\text{Variation not explained} \\ \text{by model}}$
- $SST = SSM + SSE$ (Note, you may see other acronym like $SSR$ instead of $SSE$)
- $SST$ = Sum of Squares Total (same as SSE if we use the empty model)
- $SSE$ = Sum of Squares Error (the amount of variation not explained by the model)
- $SSM$ = Sum of Squares Model (the amount of variation explained by the model)
    - We've calculated both the $SST$ and $SSE$... we can get the SSM using subtraction: $SST = SSM + SSE$
- Every point's residual is a leftover piece of variation the model didn't explain.
:::

## Putting it all together

::::{.columns}
::: {.column width="50%"}
```{r}
#| echo: true
augment(temp_model) |>
  select(count, .fitted, .resid) |> 
  head(3) |> 
  kable()
```
:::
::: {.column width="50%"}
Compute the SST, SSM, and SSE for using the linear regression model and compare it to the same values if using the mean as our model.
:::
::::


## Properties of least squares regression {.smaller}

::: incremental
-   Passes through the center of mass point, the coordinates corresponding to average $X$ and average $Y$: $\hat{\beta}_0 = \bar{Y} - \hat{\beta}_1\bar{X}$
-   Slope has same sign as the correlation coefficient: $\hat{\beta}_1 = r \frac{s_Y}{s_X}$
    + $r$: correlation coefficient
    + $s_Y, s_X$: sample standard deviations of $X$ and $Y$
-   Sum of the residuals is zero: $\sum_{i = 1}^n e_i \approx 0$ (sound familiar?)
    +   Intuition: Residuals are "balanced"
-   The residuals and $X$ values are uncorrelated
:::

## Partitioning Variability (ANOVA)

:::{.incremental}
-   $y_i - \bar{y} = (\hat{y}_i - \bar{y}) + (y_i-\hat{y}_i)$
-   Square and sum: $\sum(y_i-\bar{y})^2 = \sum(\hat{y} - \bar{y})^2 + \sum(y-\hat{y})^2$
-   $\substack{\text{Sum of squares} \\ \text{Total}} = \substack{\text{Sum of squares} \\ \text{model}} + \substack{\text{Sum of squares} \\ \text{error}}$
-   $SSTotal = SSModel + SSE$
-   $SST = SSM + SSE$
-    This process of partitioning variability is called **Analysis of Variance (ANOVA)**: Analysis of Variance
  -   Involves a whole different way of thinking about modeling and experiemental design
:::

## Residuals in R

- `augment` function takes a model and a dataset and computes the residuals (`.resid`), fitted values (`.fitted`), and many other useful quantities

```{r}
augment(temp_model, fall) |> 
  kable() |> 
  # ignore below this line
  kable_styling() |>
  scroll_box(width = "1000px", height = "300px")
```

## ANOVA in R

- `supernova` from the `supernova` package takes a linear model and decomposes the variation into it's $SSM$ and $SSE$ (look at the `sumsq` column)... which do we think is which? 
- Calculate the SST based on this information...

```{r}
temp_model |>
  supernova()
```


# Evaluating and Comparing Models

## A different model

What if we wanted to try a different model...

```{r}
workingday_model <- lm(count ~ workingday, data = fall)
workingday_model |> tidy() |> kable()
```

Is this Working Day model better?

## Which model explains more variation

```{r}
#| echo: false

p1 <- gf_point(count ~ temp_orig, data = fall) |> 
  gf_model(temp_model) |> 
  gf_labs(y = "Bike Rentals", x = "Temperature", title = "Temperature Model")
p2 <- gf_jitter(count ~ workingday, data = fall) |> 
  gf_model(workingday_model) |> 
  gf_labs(y = "Bike Rentals", title = "Working Day Model")

p1 | p2
```

## $R^2$: Percent Reduction in Error (PRE)

- Our goal is to maximize the variation explained by our model (minimize the leftover error for our model)
- It makes sense to ask... "How much has our model reduced the Total Error By":
$$
PRE = R^2 = \frac{SSM}{SST}
$$
- We call this quantity the **Precent Reduction in Error (PRE)**... 
- More commonly it is simply referred to as $R^2$

## Back to our examples {.smaller}

::::{.columns}
:::{.column}
```{r}
temp_model |> supernova()
```
:::
:::{.column}
```{r}

workingday_model |> 
  supernova()
```
:::
::::

::: question
1. Which model has a better $R^2$? 
2. What is a less common but more descriptive name of $R^2$?
3. Notice that I'm using a different model (`temp_fit`) instead of `temp_model`. `temp_fit` is estimated on the full data (i.e. all four seasons). Why do you think I didn't just use `temp_model` for the first model?
:::

## Recall: Correlation Coefficient

-   The **correlation coefficient**, $r$, is a number between -1 and +1 that measures how strong the linear relationship between two variables $x$ and $y$ is.

$$
r = \frac{\sum(x_i - \bar{x})(y_i-\bar{y})}{\sqrt{\sum(x_i-\bar{x})^2\sum(y_i-\bar{y})^2}}
= \frac{\text{Covariance of X and Y}}{(SD of X)(SD of Y)}
$$

## Interesting Property of $R^2$ {.smaller}

-   **R-squared** AKA $R^2$ AKA **PRE** AKA the **Coefficient of Determination** is not just the proportion of variability in the outcome explained by the regression model it is also the correlation between our *fitted* and *actual values* squared...
$$
R^2 = Cor(y, \hat{y})^2
$$

::: question
1. What should be the minimum and maximum value of $R^2$ based on this information?
2. Do we want $R^2$ to be big or small?
:::

## $R^2$

::: incremental
-   Ranges between 0 (terrible predictor) and 1 (perfect predictor)
-   Has no units
-   Calculate with `rsq()` from `yardstick` package from earlier using the augmented data or `supernova`:
```{r}
#| echo: true

temp_aug <- augment(temp_model)

rsq(temp_aug, truth = count, estimate = .fitted) |> kable()
```
:::

## Interpreting $R^2$ {.smaller}

```{r}
#| echo: false
temp_model_rsq <- round(rsq(temp_aug, truth = count, estimate = .fitted) |> pull(.estimate) * 100, 1)
```

::: poll
🗳️ **Discussion**

::: midi
The $R^2$ of the model for `count` from `temp_orig` is `r temp_model_rsq`%. Which of the following is the correct interpretation of this value?
:::

1.  `temp_orig` correctly predicts `r temp_model_rsq`% of `count`.
2.  `r temp_model_rsq`% of the variability in `count` can be explained by `temp_orig`.
3.  `r temp_model_rsq`% of the variability in `temp_orig` can be explained by `count`.
4.  `r temp_model_rsq`% of the time `count` can be predicted by `temp_orig`.
:::

## RMSE {.smaller}

::: incremental
-   **Root mean square error, RMSE**: A measure of the average error (average difference between observed and predicted values of the outcome)
$$
RMSE = \sqrt{\frac{\sum_{i = 1}^n (y_i - \hat{y}_i)^2}{n}}
$$
    + Sometimes people just care about numerator (SSE) or version without the square-root (MSE)
    + Sometimes the denominator may something other than $n$ like $n-1$ or $n-k$ instead
-   Interpretation (kind of): how much does my model miss by, on average.
:::

::: question
1. What should be the minimum and maximum of RMSE? [0 and infinity]{.fragment}
2. What units are RMSE in? [the same units as the response variable]{.fragment}
:::

## RMSE in R {.smaller}

```{r}
#| echo: true

yardstick::rmse(temp_aug, truth = count, estimate = .fitted)
```

- What's with the `yardstick::` before `rmse`? 
- Sometimes when you load a lot of packages... multiple packages have functions with the same name...
- In this case I'm using two packages `yardstick` and a different one called `Metrics` that both have `rmse`.
- R will try to just guess which one you want, but if it guesses wrong, you'll get an error.
- Be writing `yardstick::rmse` I'm saying "Use the `rmse` function from `yardstick`"
- This will likely not be an issue for you, if it is... that is how you solve it!


# Back to your coffee truck models

## Your three models {.smaller}

Recall from Monday, you fit three models to your own coffee truck data.

::: question
Now that you've seen residuals, SSE, $R^2$, and RMSE — which of your three models do you expect to win? Has your prediction from earlier changed?
:::

## Work time

::: appex
📋 [AE 06-07 - Comparing Models with the Coffee Truck](../ae/ae-06-07-coffee-truck-eval.qmd)
:::

Complete the remainder of the exercises.

## Quantitative vs. categorical price {.smaller}

::: incremental
- `fit_cat`:  (`Price` as the explanatory variable) fits a separate mean for each of the 4 price levels — it has more "freedom" to match the data.
- `fit_num` (`PriceNum` as the explanatory variable) forces those 4 means onto a single straight line.
- Because a line is a *special case* of "4 free means," `fit_cat`'s $R^2$ can never be smaller than `fit_num`'s $R^2$.
- So why not always use the categorical version?
    - It can't predict at a price it hasn't seen, like \$3.50.
    - It has no single "per dollar" slope to interpret or report.
    - It uses more parameters to explain (possibly) only a little more variation
:::

## Using the word "Good"

-   There is no such thing as a "Good" $R^2$ or, especially, RMSE without **context**.
-   Whether your model is a "Good" model depends on many things:
    +   What are you using your model for?
    +   How good are other models?
    +   **What else was varying in the data you fit it to?**

## Recap {.smaller}

-   A **residual** is the leftover gap between an observed value and the model's prediction: $e_i = y_i - \hat{y}_i$.
-   The **least squares method** picks $\hat{\beta}_0, \hat{\beta}_1$ to minimize $\sum e_i^2$.
-   Can decompose total variation (SST) into variation explained by the model (SSM) and leftover variation (SSE).
-   Two metrics for evaluating and comparing models:
    +   $R^2$: What proportion of the variation in the response variable is explained by the model?
    +   $RMSE$: How far does my model miss by, on average?
-   $R^2$ depends on the data it was fit to, not just the true relationship — so it can't be compared across data sets that vary different things.
