---
title: "Collecting the Islands Data"
subtitle: "Carrying out our stratified cluster sample"
author: "Prof. Eric Friedlander"
footer: "[🔗 MATH 2025 - Fall 2026 - Schedule](https://ericfriedlander.github.io/math2025-fa26/schedule)"
logo: "../images/logo.png"
format:
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk:
    R.options:
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)
library(tidyverse)   # for data wrangling

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

## Last time {.smaller}

::: {.incremental style="font-size: 0.9em"}
- A **population** is who we care about, a **sample** is who we observe, and inference uses a statistic to learn about a parameter.
- Random sampling lets us generalize, random assignment lets us conclude causation.
- We **designed** a stratified cluster sample of the Islands to estimate life expectancy, but didn't get to collect it.
:::

. . .

::: question
Today: carry out that sample, then think about who ends up **in** it. **What is the life expectancy on the Island? Is it different for men and women?**
:::

## Topics

- Carry out our **stratified cluster sample** on the Islands.
- Think about **who is (and isn't)** in our sample.
- **Clean** the pooled class data together.
- Get a head start on **HW 05**.

## Activity

::: appex
📋 [AE 09 - Collecting the Islands Data](../ae/ae-09-islands-collection.qmd)
:::

Open and upload your QMD file to RStudio.

# Collecting the data

## Our sampling plan {.smaller}

You will adapt the stratified cluster design:

1. Different Islands might have drastically different life expectancies so let's make sure we sample individuals from each island (i.e. stratify by island)
2. Individual villages within each Island may also have some variation in life expectancy. However, it will be too difficult to sample EVERY village so we'll treat these as clusters.
3. Each person in this class will be randomly assigned one village.
4. Within each village, you will record data on 10 randomly collected islander.

## Overall Sampling Plan

Multistage sample:

- Stage 1: Stratify by island
- Stage 2: Cluster by village
- Stage 3: Simple random sample from those who died between 375 and 399

## Assigning Islands

Navigate to [this shared Excel sheet](https://yotescollegeofidaho-my.sharepoint.com/:x:/g/personal/efriedlander_collegeofidaho_edu/IQB2Q0ijZYPzQYuumjlk_x09AQaTTMpIIkDa73TeNCMKJiI?e=16bSqH) and click on the "Village Assignment" worksheet.

- To randomly assign you to Islands
  - Write down everyone's name
  - Randomize order of Islands
  - Assign name to Island in order

::: question
Discussion: Should we use the same number of villages from each island?
:::

## Count the number of dead

- Create a new worksheet (the + button at the bottom), rename it with your village's name, and then copy and paste the list of islanders into that sheet

::: question
- How many islanders are in your village's list?
- Should we sample the same number of islanders from each village?
:::

## Choosing your 10 islanders {.smaller}

Randomly select 10 of them:

- In your Quarto document, input your birthday and the total number of villages and run the following code to choose 10 random row numbers:

```{r}
#| eval: false

birthday <- # put your birthday here in MMDDYYYY format (e.g. 12181988)
set.seed(birthday)
number_of_villagers <- # put your total number of islanders here
sample(1:number_of_villagers, 10)
```

## Entering your data {.smaller}

Use the numbers to identify which islanders to sample, and record your name, the island, the village, the villager's name, sex, birth year, and death year in the **MASTER DATASET** worksheet in that same shared Excel sheet.

::: callout-important
Only add rows, never delete or overwrite anyone else's. Enter `sex` as exactly `Female` or `Male`, and enter every value the same way each time.
:::

## Activity

Complete the AE (getting oriented, collecting your sample, entering your data, and thinking about who is in the sample).

```{r}
#| echo: false
countdown(minutes = 20)
```

# Thinking about who is in the sample

## Who is in our sample? {.smaller}

We are only sampling islanders who died between Island years 375 and 399.

::: question
- Who **cannot** possibly appear in this sample?
- Why might this make our data a poor stand-in for the life expectancy of a *newborn* today?
- Name one way our sampling plan is better than just picking the first 10 islanders we come across.
:::

## Time permitting {.smaller}

::: question
1. We are selecting the same number of villages from each island and the same number of individuals from each village... could this bias our results?
2. We aren't stratifying by sex. Perhaps it would've been a good idea to do so... Can you think of a situation in which this isn't a good idea? Hint: it may help if you imagine that the islanders are an alien species different than humans.
:::

# Cleaning the class data

## Real data are messy {.smaller}

Everyone typed in their own rows, so expect some surprises.

::: incremental
- Is `sex` spelled the same way every time?
- Are there duplicate rows, missing values, or impossible ages?
- Does every column have the type we expect (numbers stored as numbers)?
:::

. . .

We create `age_at_death = death_year - birth_year`, and then check the values before we trust them.

## Activity {.smaller}

::: appex
📋 [AE 09 - Collecting the Islands Data](../ae/ae-09-islands-collection.qmd)
:::

Complete Exercise 6 (load, inspect, and clean the pooled data together). If you finish early, model the difference between men and women in the time-permitting section.

```{r}
#| echo: false
countdown(minutes = 15)
```

## Finished early? {.smaller}

::: appex
📝 [HW 05 - Designing Studies and Modeling the Islands Data](../hw/hw-05.qmd)
:::

Part 2 of HW 05 is about study design and doesn't use any data, so you can start it now if you're done entering your islanders. Part 1 has you model the pooled data, which we'll have after this class.

## Recap {.smaller}

-   We carried out a **stratified cluster sample** of the Islands: each of us drew 10 islanders who died in Island years 375-399 from our assigned village.
-   The mean age at death among islanders who have died is a stand-in for life expectancy, and it is biased because of *who can appear* in the data.
-   Hand-entered data need **cleaning** before we can trust them: check spelling, duplicates, missing values, and impossible values.
-   **Next time:** we explore the pooled class data and ask whether sex explains **more** variation in age at death than we'd expect by chance.
