---
title: "Modeling Life Expectancy"
subtitle: "Analyzing the data we collected on the Islands"
author: "Prof. Eric Friedlander"
footer: "[🔗 MATH 2025 - Fall 2026 - Schedule](https://ericfriedlander.github.io/math2025-fa26/schedule)"
logo: "../images/logo.png"
format:
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk:
    R.options:
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

# load packages
library(tidyverse)   # for data wrangling and visualization
library(ggformula)   # for plotting using formulas
library(broom)       # for formatting model output
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(kableExtra)  # also for pretty tables

# MADE-UP preview data: a pretend Island population, used only so these
# slides render without the class data. Real data comes
# from the AE.
set.seed(2025)
n_pop <- 5000
island_pop <- tibble(
  sex = sample(c("Female", "Male"), n_pop, replace = TRUE),
  age_at_death = round(pmin(pmax(
    rnorm(n_pop, mean = if_else(sex == "Female", 72, 69), sd = 16),
    0), 120))
)
islands <- slice_sample(island_pop, n = 250)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

## Last time {.smaller}

::: {.incremental style="font-size: 0.9em"}
- A **population** is who we care about, a **sample** is who we observe, and inference uses a statistic to learn about a parameter.
- Random sampling lets us generalize, random assignment lets us conclude causation.
- We took a **stratified cluster sample** of the Islands: each of us drew 10 islanders who died in Island years 375-399 from our assigned village.
:::

. . .

::: question
We now have one pooled data set. **What is the life expectancy on the Island? Is it different for men and women?**
:::

## Topics

- Load and **clean** the pooled data that the class entered by hand.
- Estimate life expectancy and fit a model predicting age at death from sex.
- Critique what our sample can, and cannot, tell us.
- See **sampling variability** first-hand: same population, different samples, different slopes.

# Loading and cleaning the class data

## Real data are messy {.smaller}

Everyone typed in their own rows, so expect some surprises.

::: incremental
- Is `sex` spelled the same way every time?
- Are there duplicate rows, missing values, or impossible ages?
- Does every column have the type we expect (numbers stored as numbers)?
:::

. . .

We create `age_at_death = death_year - birth_year`, and then check the values before we trust them.

## Activity {.smaller}

::: appex
📋 [AE 09 - Analyzing the Islands Data](../ae/ae-09-islands-analysis.qmd)
:::

Complete Exercise 1 (load, inspect, and clean the pooled data).

```{r}
#| echo: false
countdown(minutes = 15)
```

# Modeling life expectancy from sex

## Our pooled data {.smaller}

::: callout-note
The output on these slides is from a **made-up** island so the slides render without the class data. Yours will look different.
:::

```{r}
islands |>
  head(5) |>
  kable()
```

::: question
What are the observational units? What are the variables, and which is the response?
:::

## Exploring age at death {.smaller}

```{r}
#| out.width: "70%"

gf_boxplot(age_at_death ~ sex, data = islands, fill = ~ sex, show.legend = FALSE) |>
  gf_labs(x = "Sex", y = "Age at death (years)")
```

## Fitting the model {.smaller}

```{r}
#| echo: true

le_fit <- lm(age_at_death ~ sex, data = islands)

tidy(le_fit)
```

```{r}
#| echo: false

b0 <- coef(le_fit)[["(Intercept)"]]
b1 <- coef(le_fit)[["sexMale"]]
```

$$
\widehat{\text{age at death}} = `r round(b0, 2)` `r if_else(b1 < 0, "-", "+")` `r round(abs(b1), 2)` \times \text{sexMale}
$$

::: question
- What is the reference level? How can you tell?
- What does the intercept mean? What does the slope mean?
:::

## Interpreting the model {.smaller}

- **Intercept:** the mean age at death for the reference group (`Female`): `r round(b0, 2)` years.
- **Slope:** the *difference in means*. Men in our sample die `r round(abs(b1), 2)` years `r if_else(b1 < 0, "earlier", "later")`, on average, than women.

. . .

```{r}
#| echo: true

islands |>
  group_by(sex) |>
  summarize(mean_age = mean(age_at_death))
```

This is the same categorical-predictor model from Lesson 5.

## Work time {.smaller}

::: appex
📋 [AE 09 - Analyzing the Islands Data](../ae/ae-09-islands-analysis.qmd)
:::

Complete Exercises 2-5 (explore, fit, interpret, and critique).

```{r}
#| echo: false
countdown(minutes = 15)
```

# Is the difference between men and women real?

## The question {.smaller}

Our model says men and women differ by about `r round(abs(b1), 2)` years in our sample.

::: question
**Is the difference between men and women real?**
:::

. . .

*What could "real" mean here?*

## Same population, different samples {.smaller}

In our made-up world we *know* the population, so let's take four more samples of the same size.

```{r}
#| echo: true

set.seed(101)

map_dfr(1:4, \(i) {
  island_pop |>
    slice_sample(n = 250) |>
    lm(age_at_death ~ sex, data = _) |>
    tidy() |>
    filter(term == "sexMale") |>
    mutate(sample = i)
}) |>
  select(sample, estimate) |>
  kable(digits = 2)
```

::: question
Same island, same model, same sample size. Four different slopes. Which one is "right"?
:::

## Two competing explanations {.smaller}

If a sample slope $\hat{\beta}_1$ is not zero, there are (at least) two reasons why:

::: incremental
1. There **is** a real difference in the population: $\beta_1 \neq 0$.
2. There is no real difference, and we just got a **lucky (or unlucky) draw**: $\beta_1 = 0$ but $\hat{\beta}_1 \neq 0$.
:::

. . .

Sampling variability means a sample statistic won't equal the population parameter, so **the data alone can't tell us which explanation is right.** We need a way to measure how much a slope is *expected* to bounce around.

. . .

::: question
Look at the four slopes on the previous slide. How big would the slope have to be before *you'd* be convinced?
:::

## Back to the AE {.smaller}

::: appex
📋 [AE 09 - Analyzing the Islands Data](../ae/ae-09-islands-analysis.qmd)
:::

Complete Exercises 6-7: refit the model using only the 10 islanders *you* sampled, compare slopes on the board, and decide how confident you are.

```{r}
#| echo: false
countdown(minutes = 10)
```

## Recap {.smaller}

-   Hand-entered data need **cleaning** before we can trust them: check spelling, duplicates, missing values, and impossible values.
-   The mean age at death among islanders who have died is a stand-in for life expectancy, and it is biased because of *who can appear* in the data.
-   A regression with a categorical predictor estimates a **difference in means**, but a different sample gives a different $\hat{\beta}_1$.
-   **Next time:** how do we measure how much a slope should bounce around by chance?
