---
title: "AE 03: Dr. F's Coffee"
subtitle: "Simple linear regression"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        </style>
---

::: callout-important
-   Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-03"

-   Go to the [Canvas](https://cofi.instructure.com/courses/18743) and locate the `AE-03` assignment. That link should take you directly there.

-   Upload the `ae-03-coffee.qmd` file into the folder you just created. There is no separate data file: today's data loads directly from the internet, so you'll need to be online when you render.
:::

```{r load-packages}
#| message: false
library(tidyverse)
library(broom)
library(ggformula)
library(janitor)
```

## Data

Our data set is Dr. F's own coffee log: every cup of coffee he's brewed at home since January 2026, along with the beans, brewing method, and his own ratings of how the cup turned out. Each row is a single cup of coffee.

We will focus on the following variables in the analysis:

-   `vibes_rating`: Dr. F's overall rating of the cup, on a scale from 0 to 10.
-   `body`: sub-score for the cup's body (mouthfeel), also on a 0 to 10 scale.
-   `roast`: the roast level of the beans (`Light`, `Light-Medium`, `Medium`, or `Dark`).

::: callout-note
There's also an `average` column, calculated from `body` and four other flavor sub-scores. Since it's essentially a summary of variables we're already using, we won't use it here, same as in HW 01.
:::

## Exercises

::: {.exercise}
The following code will load the data.

```{r load-data}
#| message: false
coffee <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTI_1K-jHHdbpi11aB5tX3Avr2eRXULfLg9BoveWhId4soH1BAnIJWuu0CiqO7YiP1TVoqsGi6kLbli/pub?gid=0&single=true&output=csv") |>
  clean_names()
```
:::

::: {.answer}

:::

::: {.exercise}
Below is code to create visualizations of the distributions of `vibes_rating` and `body`. Create a bar chart of `roast` and store it in a variable called `p3`. Make sure to include appropriate labels. Finally, uncomment the last line of code in this chunk. What do you think this last line does? Comment on what you see.

```{r coffee-viz}
#| warning: false
#| message: false

p1 <- gf_histogram(~ vibes_rating, data = coffee, binwidth = 0.5) |>
  gf_labs(x = "Vibes rating")

p2 <- gf_histogram(~ body, data = coffee, binwidth = 1) |>
  gf_labs(x = "Body")

# (p1 | p2) / p3
```
:::

::: {.answer}

```{r}

```

:::

::: {.exercise}
In the raw data, `roast` is a character variable, and its levels won't sort in a sensible order on their own (alphabetical would put `Dark` first). Recode `roast` to make it a factor with levels in order from lightest to darkest roast. While you're at it, take a look at the `origin` column: fix any typos you find that would cause the same country to be counted as two different categories.

```{r data-prep}
# add code developed during livecoding here
```
:::

::: {.answer}

:::

::: {.exercise}
We want to evaluate whether the relationship between `body` and `vibes_rating` is the same for each roast level. To answer this question, first create a scatter plot of `vibes_rating` vs. `body`, faceted by `roast`. Since `body` only takes on a handful of integer values and lots of points will land on top of each other, use `gf_jitter` instead of `gf_point`.

```{r facet-plot}
# add code developed during livecoding here
```
:::

::: {.answer}

```{r}

```

:::

::: {.exercise}
Does the relationship between `body` and `vibes_rating` look about the same across the four roast levels, or different? Which panel(s), if any, should we be cautious about drawing conclusions from, and why?
:::

::: {.answer}

:::

::: {.exercise}
Use `filter` and `arrange` to find the 5 lowest-rated cups in the data set. Do they have anything in common (roast, origin, brewer, a particular sub-score)?

```{r filter-lowest}
# add code developed during livecoding here
```
:::

::: {.answer}

```{r}

```

:::

::: {.exercise}
Fit a linear model for predicting `vibes_rating` from `body`, using all of the cups (no filtering by roast). Use `tidy()` to format the output.

```{r fit-model}
# add code developed during livecoding here
```
:::

::: {.answer}

```{r}

```

:::

::: {.exercise}
Use the output to write out the estimated regression equation.
:::

::: {.answer}

:::

::: {.exercise}
Interpret the slope in the context of the data.
:::

::: {.answer}

:::

::: {.exercise}
Interpret the intercept in the context of the data. Then explain why this intercept isn't really meaningful here: is `body = 0` a value we actually observe in this data set?
:::

::: {.answer}

:::

::: {.exercise}
Create a visualization of the model we just fit.

```{r visualize-model}
# add code developed during livecoding here
```
:::

::: {.answer}

```{r}

```

:::

::: callout-important
To submit the AE:

-   Render the document to produce the HTML with all of your work from today's class.
-   The driver for your group should upload your .qmd and .html files to the Canvas assignment.
:::
