---
title: "HW 05: Modeling the Islands Data and Reading Two Studies"
author: "Insert Name Here"
subtitle: "Your first Islands models, plus generalizability and causation in two published studies"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

```{r}
#| include: FALSE
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```


# Introduction

In lecture, we learned that how data are collected determines what we can conclude from them. In this HW you'll apply what you've learned about **study design** to two recent published studies. An optional extra-credit part lets you build and interpret regression models using the class's pooled Islands data from AE 09.

**Part 1** is about reading two studies and needs no data, you can start it as soon as we finish collecting our sample in Lesson 9. **Part 2** is extra credit and uses the pooled class data.

::: callout-note
This is an **individual** assignment, but you're welcome to talk through ideas with classmates. Your code and writing should be your own.
:::

## Learning goals

By the end of this homework, you will...

- Classify studies as observational or experimental, and say what each design can and cannot conclude
- Read key parts of a research paper to identify its design, its population, and its limitations
- Design an experiment to address a question that a published observational study can't fully answer
- Explain how random sampling and random assignment separately determine generalizability and the ability to establish causation
- Identify confounding variables and threats to the validity of a study
- (Extra credit) Fit, interpret, and evaluate simple linear regression models with categorical and quantitative predictors
- (Extra credit) Describe the scope of inference for a set of models

# Getting Started

### Log in to RStudio

- Go to [RStudio](https://rstudio.collegeofidaho.edu/) and login with your College of Idaho Email and Password.

- Use the `hw` folder you made earlier in the semester. Create a subfolder called `hw05` to store this homework.

- Log into [Canvas](https://cofi.instructure.com/courses/18743/assignments/244108), navigate to HW-05, and download the `hw-05.qmd` file. Upload it into the folder you just made.

## R and RStudio

See the [R resources](https://ericfriedlander.github.io/math2025-fa26/computing-r-resources.html) page for a guide to the RStudio IDE and Quarto file components.

# Packages

We will use the following packages in today's HW.

```{r load-packages}
#| message: false
library(tidyverse)
library(mosaic)
library(ggformula)
library(broom)
library(coursekata)
library(yardstick)
```

# Background reading

(Optional reference for Part 1.) Read the chapter on study design in *Introduction to Modern Statistics*: [Data collection](https://openintrostat.github.io/ims/data-design.html). Pay special attention to the sections on observational studies vs. experiments, confounding, and sampling.

# Exercises

::: callout-tip
Write all code and narrative in your Quarto file. Write all narrative in complete sentences. Throughout the assignment, you should periodically **render** your Quarto document to produce the updated HTML file.

Make sure we can read all of your code in your HTML file. This means you will need to break up long lines of code. One way to help avoid long lines of code is to start a new line after every pipe (`|>`).

In your `hw-05.qmd` document you'll see that each task is wrapped in an `exercise` box, with an empty `answer` box below it for you to fill in your response. Use this format for the remaining exercises.
:::

# Part 1: Reading two studies

::: callout-tip
Part 1 doesn't require any data. You can work on it as soon as we finish collecting our sample in Lesson 9, even before we've pooled and cleaned it.
:::

You do **not** need to read these papers in full. Read only the parts listed below (about five pages in total), and answer the questions. Note that there are many parts of these papers that you will not understand and that's ok. Try to focus on understanding the experimental design of each study and what the impact is on the potential conclusions of each.

- **Paper A:** Grimon and Mills, ["Better Together? A Field Experiment on Human-Algorithm Interaction in Child Protection"](https://arxiv.org/pdf/2502.08501). Child protective services workers in one county, referred to as the "trial county" in the paper, decide which reported families to investigate. Some were given access to a computer risk score, and researchers tracked what happened to the children afterward. *Read:* the abstract, the Introduction on pages 1-3, and Section 3.2 ("Sample Characteristics") and Section 3.3 ("Experimental Design") on pages 10-11.
- **Paper B:** Goodman and Nitkin, ["Kumon Up: Private Tutoring Centers and Student Achievement"](https://www.nber.org/system/files/working_papers/w35671/w35671.pdf). Chains like Kumon and Sylvan have opened thousands of tutoring centers. The authors ask who uses them and whether they raise test scores. *Read:* the abstract and the Introduction, pages 1-4.

::: exercise
**Paper A.** In the Introduction, the authors describe their study as a "large-scale field experiment ($N = 4{,}681$)".

- What was **randomized**? At what level (child, household, or social worker), and why do you think the researchers randomized at that level?
- What are the **treatment** and **control** conditions?
- What is the main **outcome** they use to judge whether the algorithm helped?
- In the Introduction (page 3), the authors say an algorithm-only condition (with no human decision-maker) was "legally and practically infeasible." Why might a researcher be unable to run the experiment they would ideally want?
:::

::: answer
:::

::: exercise
**Paper A.** Workers were not required to use the algorithm. The Introduction says workers "consulted the algorithm in at least 73 percent of decisions."

Suppose a researcher ignored the randomization and simply compared outcomes for families where workers *did* look at the score to families where they *didn't*.

- Explain why this would be an **observational** comparison, even though it happens inside an experiment.
- Name one plausible **confounding variable**, and explain how it could make the algorithm look better (or worse) than it really is.
- How does the randomization protect the researchers from this problem?
:::

::: answer
:::

::: exercise
**Paper A.** Read the county description in Section 3.3 (page 11): the trial county "ranks above the national average in socioeconomic status," is "predominantly white (92%)," and "uses group decision-making."

- Did the researchers take a **random sample** of child protection cases from across the country? How can you tell?
- To what population can the results **safely generalize**?
- Name one type of place where you would hesitate to expect the same result, and explain why.
- Does the lack of random sampling weaken the claim that the algorithm *caused* the improvement? Why or why not?
:::

::: answer
:::

::: exercise
**Paper B.** Read the second and third paragraphs of the Introduction, and the abstract.

- Is this an **observational study** or an **experiment**? What tells you?
- In your own words, what is being compared? (Hint: the authors compare districts that gain a first chain center between 2010 and 2019 to districts that never do.)
- The abstract says "Asian and White households \[are\] five and two times as likely to purchase tutoring as Black and Hispanic households." Is this an *association* or a *causal* claim? Explain.
- Who decides where a chain tutoring center opens? Explain why this makes it hard to tell whether the centers *cause* higher test scores.
:::

::: answer
:::

::: exercise
**Paper B.** Districts that got their first chain center could have been on a path of rising scores even *without* the center. On page 2-3 of the Introduction, the authors report "flat pre-trends": before a center opened, scores in treated districts were not already rising faster than in other districts.

- Explain why flat pre-trends make the causal claim more believable, but do not fully settle it. Name one thing that could still go wrong.
- Give one **confounding variable** (something that changes at the same time as a center opening) that the design might not catch.
:::

::: answer
:::

::: exercise
**Paper B.** Read the paragraph on page 4 beginning "How large are these effects on the students who actually attend these centers?"

- The district-wide effect is about 0.03 standard deviations, but the implied effect on students who attend is about 0.25 standard deviations or more. Why are these two numbers different? What **population** does each one describe?
- The results are much larger for White and Asian students than for Black and Hispanic students. Does this tell us that tutoring *works differently* for different groups? What else might explain the difference?
- The cover page says NBER working papers "have not been peer-reviewed." How should that change how you treat the paper's claims?
:::

::: answer
:::

::: exercise
**Comparing the papers.** Two design choices control what a study can conclude: **random sampling** (can we generalize?) and **random assignment** (can we claim causation?).

Fill in the table for each paper, and briefly justify each answer with something specific from the reading.

|   | Random sampling of the population? | Random assignment? | Can generalize? | Can claim causation? |
|:--------------|:--------------|:--------------|:--------------|:--------------|
| Paper A (algorithm) |  |  |  |  |
| Paper B (tutoring) |  |  |  |  |

Then, in two or three sentences, say which paper you trust more for the claim "this intervention *causes* better outcomes," and which for "this result holds broadly across the country." Do they give the same answer?
:::

::: answer
:::

::: exercise
Suppose you had unlimited resources and wanted a more convincing answer to Paper B's question: *does opening a tutoring center raise test scores?*

- Design an **experiment** to answer it. Say what you would randomize, who the participants are, how you would assign them, and what you would measure. (For inspiration, the authors mention a randomized trial of Kumon in Bangladeshi schools.)
- What could your experiment tell you that Paper B cannot?
- What could Paper B tell you that your experiment could not? (Think about who, where, and for how long.)
:::

::: answer
:::

# Part 2: Extra credit, modeling the class data

::: callout-important
This part is **optional extra credit**, worth up to 3.5 points. It needs the pooled Islands data from AE 09.
:::

::: exercise
Read the pooled data into R using the code below. **Clean the data yourself.** Everyone typed in their own rows, so check for problems: is `sex` spelled consistently? Are there missing values, duplicated rows, columns stored as the wrong type, or impossible ages? Then compute `age_at_death` as `death_year - birth_year`. Note that we did this together in class and you can copy and paste from the activity. However, I encourage you to do as much as you can by yourself.

Save the cleaned result as `islands`.
:::

::: answer
```{r read-data}
#| message: false

islands_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTonqSnj8t3ZgjDzqCuq-P8bE-1OuHmwgBI0kF7eip7gkSmrJQGYRVwFtehtVS4FM6zwNAHUl1He29A/pub?output=csv")
```

```{r clean-data}

```
:::

::: exercise
Fit a linear model called `le_fit` predicting `age_at_death` from `sex`. Use `tidy()` to format the output.

- What is the **reference level** in your model? How can you tell from the output alone?
- Write out the estimated regression equation using R's coefficient names.
:::

::: answer
```{r fit-le}

```
:::

::: exercise
Interpret the intercept and the slope of `le_fit` in the context of the data.
:::

::: answer
```{r check-le}

```
:::

::: exercise
Evaluate `le_fit`.

- Use `augment()` to get fitted values and save the result to a new data frame.
- Use the `rsq()` and `rmse()` functions from `yardstick` to compute $R^2$ and RMSE.
- Write one sentence interpreting $R^2$ and one sentence interpreting RMSE in the context of your data.
:::

::: answer
```{r evaluate-le}

```
:::

::: callout-tip
This is a good place to render your document.
:::

::: exercise
Now fit a second model, `island_fit`, predicting `age_at_death` from `island` (a categorical predictor with more than two levels). Use `tidy()` to display the output.

- What is the reference level?
- Interpret **every** coefficient in the model.
- Compare the $R^2$ of `island_fit` to the $R^2$ of `le_fit`. Which predictor, sex or island, explains more variation in age at death in our sample?
:::

::: answer
```{r fit-island}

```
:::

::: exercise
Finally, fit a third model, `birth_fit`, predicting `age_at_death` from `birth_year` (a quantitative predictor).

- Make a scatterplot of `age_at_death` vs. `birth_year` with the fitted line overlaid.
- Interpret the slope. On its face, is this a strong relationship?
- Now think about *how* we collected the data: we only sampled islanders who **died between Island years 375 and 399**. Explain why this makes the relationship between `birth_year` and `age_at_death` largely an **artifact of our sampling window**. Hint: think about the sampling window.
:::

::: answer
```{r fit-birth}

```
:::

::: exercise
Think about the **scope of inference** for `le_fit`, `island_fit`, and `birth_fit`.

- To what population do these models' results generalize? Is it the whole Island, all islanders who have ever lived, or something narrower?
- Can any of these models support a **causal** claim about its predictor? Why or why not?
- What would have to be true about how the data were collected for a causal claim to be justified? Would that even be possible for `island_fit`?
:::

::: answer
:::

::: callout-tip
You're done and ready to submit your work! Render your document and upload the .qmd and .html files to Canvas.
:::

# Grading (30 pts + 3.5 Extra Credit)

<br>

| Component                        | Points |
|:---------------------------------|:-------|
| Ex 1                             | 3      |
| Ex 2                             | 3      |
| Ex 3                             | 3      |
| Ex 4                             | 3      |
| Ex 5                             | 3      |
| Ex 6                             | 3      |
| Ex 7                             | 5      |
| Ex 8                             | 5      |
| Extra credit (Ex 9-15, 0.5 each) | +3.5   |
| Grammar & Writing                | 1[^1]  |
| Workflow & formatting            | 1[^2]  |

[^1]: The "Grammar & Writing" grade is decided based on your grammar and writing. This is typically decided by choosing one of the questions and assessing the writing.

[^2]: The "Workflow & formatting" grade is to assess the reproducible workflow and document format. This includes having a neatly organized document with readable code and your name and the date in the YAML.