---
title: "HW 06: What Makes a Song Energetic?"
author: "Insert Name Here"
subtitle: "Test statistics, null distributions, and t-tests for the slope"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

```{r}
#| include: FALSE
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

# Introduction

In lecture, we learned a general recipe for inference: choose a **test statistic** that captures your research question, find its **null distribution**, and see where your data land. We saw that the null distribution of a slope is **normal**, which lets us replace shuffling with math. In this HW you'll run that whole recipe on a new question about Spotify songs.

::: callout-note
This is an **individual** assignment, but you're welcome to talk through ideas with classmates. Your code and writing should be your own.
:::

## Learning goals

By the end of this homework, you will...

- Choose a test statistic that matches a research question
- Build a null distribution by shuffling and describe its shape, center, and spread
- Use the normal and $t$ distributions to compute a p-value and relate it to the simulated p-value
- Compute and interpret a $t$ statistic by hand and sketch the p-value it corresponds to
- Describe the scope of inference of a regression analysis

# Getting Started

### Log in to RStudio

- Go to [RStudio](https://rstudio.collegeofidaho.edu/) and login with your College of Idaho Email and Password.

- Use the `hw` folder you made earlier in the semester. Create a subfolder called `hw06` to store this homework.

- Log into [Canvas](https://cofi.instructure.com/courses/18743/assignments/244332), navigate to HW-06, and download the `hw-06.qmd` and `spotify-popular.csv` files. Upload them into the folder you just made.

## R and RStudio

See the [R resources](https://ericfriedlander.github.io/math2025-fa26/computing-r-resources.html) page for a guide to the RStudio IDE and Quarto file components.

# Packages

We will use the following packages in today's HW.

```{r load-packages}
#| message: false
library(tidyverse)
library(ggformula)
library(broom)
library(knitr)
library(coursekata)  # for shuffle(), resample(), do(), b1(), pre()
library(openintro)   # for normTail()
```

# Data: Spotify

The data are the same Spotify songs from AE 11: a subset of the [Spotify Songs](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-21/readme.md) Tidy Tuesday data set, restricted to popular songs (`track_popularity >= 80`). This time our response variable is `energy`, a measure from 0.0 to 1.0 of the intensity and activity of a track. Fast, loud, noisy tracks have high energy, while a Bach prelude has low energy.

Choose **one** explanatory variable from the list below. You'll use it for the whole assignment.

| variable | description |
|:----------------------|:-----------------------------------------------|
| loudness | Overall loudness of a track in decibels (dB) |
| acousticness | Confidence from 0.0 to 1.0 that the track is acoustic |
| valence | Musical positiveness, from 0.0 (sad, angry) to 1.0 (happy, euphoric) |
| tempo | Estimated tempo in beats per minute (BPM) |
| danceability | How suitable a track is for dancing, from 0.0 to 1.0 |
| speechiness | Presence of spoken words, from 0.0 to 1.0 |
| liveness | Likelihood from 0.0 to 1.0 that the track was performed live |
| duration_min | Duration of the song in minutes |

```{r load-data}
spotify <- read_csv(here::here("data/spotify-popular.csv")) |>
  mutate(duration_min = duration_ms / 60000)
```

# Exercises

::: callout-tip
Write all code and narrative in your Quarto file. Write all narrative in complete sentences. Throughout the assignment, you should periodically **render** your Quarto document to produce the updated HTML file.

Make sure we can read all of your code in your HTML file. This means you will need to break up long lines of code. One way to help avoid long lines of code is to start a new line after every pipe (`|>`).

In your `hw-06.qmd` document you'll see that each task is wrapped in an `exercise` box, with an empty `answer` box below it for you to fill in your response. Use this format for the remaining exercises.
:::

::: exercise
Which explanatory variable did you choose? Make a visualization of `energy`, a visualization of your explanatory variable, and a scatterplot of `energy` against your explanatory variable. Include informative titles and axis labels. Describe the distribution of `energy` and the relationship you see in the scatterplot.
:::

::: answer
```{r eda}

```
:::

::: exercise
Fit a linear model predicting `energy` from your explanatory variable and save it as `energy_fit`. Use `tidy()` and `kable()` to display it. Write out the fitted equation and interpret the intercept and slope in the context of the data.
:::

::: answer
```{r fit-model}

```
:::

::: exercise
**Choosing a statistic.** For each research question below, name a test statistic that captures it (for example: a slope, a difference in group means, or PRE) and briefly explain your choice.

a.  Is there a linear relationship between your explanatory variable and `energy`?
b.  Do songs in a major key (`mode = 1`) have a different average `energy` than songs in a minor key (`mode = 0`)?
c.  Does knowing a song's `playlist_genre` (a variable with several categories) explain more of the variation in `energy` than we'd expect by chance?
:::

::: answer
:::

::: exercise
**The null distribution.** For question (a) in Exercise 3, write the null and alternative hypotheses in words. Then:

- Save your observed slope as `b1_obs`.
- Build a null distribution of the slope by shuffling `energy` 1000 times with `do(1000) * b1(shuffle(energy) ~ your_variable, data = spotify)`.
- Make a histogram of the null distribution with a vertical line at `b1_obs`.
- Describe the shape, center, and spread of the null distribution. Why is it centered where it is?
- Compute the simulation p-value. Why do we need `abs()`?
:::

::: answer
```{r null-dist}

```
:::

::: exercise
**The normal distribution.** The null distribution you built looks like a normal distribution centered at 0.

- Compute the standard deviation of your shuffled slopes and compare it to the `std.error` for your slope in `tidy(energy_fit)`.
- Use `pnorm()` with mean 0 and the SD of your shuffled slopes to compute a two-sided p-value for your observed slope. How does it compare to the simulation p-value?
- What does this tell you about whether we really need to shuffle 1000 times?
:::

::: answer
```{r normal-approx}

```
:::

::: exercise
**The** $t$ statistic. Using the output of `tidy(energy_fit)` do the following:

- Write down the correct t-statistic and interpret it: how many standard errors is your slope from the null value, and in which direction?
- Find the p-value in the output above.
- Write a conclusion in the context of the problem. How strong is the evidence against the null hypothesis?
- We use a $t$ distribution rather than a normal distribution. In one sentence, why?
:::

::: answer
```{r t-stat}

```
:::

::: exercise
Complete the following on paper and upload a picture of your work along with this assignment:

- Write down the null and alternative hypothesis in symbols.
- Compute the test statistic $T = (\hat{\beta}_1 - 0)/SE_{\hat{\beta}_1}$ and verify that it matches the output.
- Sketch a t-distribution and shade in the area corresponding to the p-value.
:::

::: callout-tip
This is a good place to render your document.
:::

::: exercise
**Scope of inference.**

- To what population do your results generalize? Think about how these songs were selected (hint: `track_popularity >= 80`).
- Was this a random sample or an experiment? Can you claim that changing your explanatory variable *causes* a change in `energy`? Why or why not?
- Name one other variable that could confound the relationship you studied.
:::

::: answer
:::

::: callout-tip
You're done and ready to submit your work! Render your document and upload the .qmd and .html files to Canvas.
:::

# Grading (28 pts + 2 pts for writing and workflow)

<br>

| Component             | Points |
|:----------------------|:-------|
| Ex 1                  | 3      |
| Ex 2                  | 3      |
| Ex 3                  | 3      |
| Ex 4                  | 4      |
| Ex 5                  | 4      |
| Ex 6                  | 4      |
| Ex 7                  | 4      |
| Ex 8                  | 3      |
| Grammar & Writing     | 1[^1]  |
| Workflow & formatting | 1[^2]  |

[^1]: The "Grammar & Writing" grade is decided based on your grammar and writing. This is typically decided by choosing one of the questions and assessing the writing.

[^2]: The "Workflow & formatting" grade is to assess the reproducible workflow and document format. This includes having a neatly organized document with readable code and your name and the date in the YAML.