---
title: "HW 04: Education & median income in US Counties"
author: "Insert Name Here"
subtitle: "Residuals, R-squared, and RMSE"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

```{r}
#| include: FALSE
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

<!-- TODO (instructor): fill in the Canvas link in "Getting Started" before posting. -->

# Introduction

In this homework, you will use simple linear regression to examine the association between the percent of adults with a bachelor's degree and the median household income for counties in the United States. This builds directly on lessons 6 and 7: you'll fit a model, compute and interpret residuals, and use $R^2$ and RMSE to evaluate and compare models.

## Learning goals

By the end of this homework, you will...

- Fit and interpret a simple linear regression model
- Compute and interpret residuals for individual observations
- Decompose total variation into variation explained by the model and leftover variation ($SST = SSM + SSE$)
- Compute and interpret $R^2$ and RMSE
- Compare two models using $R^2$ and RMSE

# Getting Started

### Log in to RStudio

- Go to [RStudio](https://rstudio.collegeofidaho.edu/) and login with your College of Idaho Email and Password.

- Use the `hw` folder you made earlier in the semester. Create a subfolder called `hw04` to store this homework, and a `data` subfolder inside of it.

- Log into [Canvas](), navigate to HW-04, and download the `hw-04.qmd` and `us-counties-sample.csv` files. Upload the `.qmd` into the folder you just made, and the `.csv` into the `data` subfolder.

## R and RStudio

See the [R resources](https://ericfriedlander.github.io/math2025-fa26/computing-r-resources.html) page for a guide to the RStudio IDE and Quarto file components.

# Packages

We will use the following packages in today's HW.

```{r load-packages}
#| message: false
library(tidyverse)
library(mosaic)
library(ggformula)
library(broom)
library(coursekata)
library(yardstick)
library(supernova)
```

# Data: US Counties

The data are from the `county_2019` data frame in the [usdata](https://openintrostat.github.io/usdata/index.html) R package. These data were originally collected in the 2019 [American Community Survey](https://www.census.gov/programs-surveys/acs) (ACS), an annual survey conducted by the United States Census Bureau that collects demographics and other information from a sample of households in the United States. The data in `county_2019` are county-level statistics from the ACS.

The primary variables you will use are the following:

-   `bachelors`: Percent of population 25 years old and older that earned a Bachelor's degree or higher
-   `median_household_income`: Median household income in US dollars
-   `household_has_computer`: Percent of households that have a desktop or laptop computer

[Click here](https://openintrostat.github.io/usdata/reference/county_2019.html) for the full codebook for the `county_2019` data set.

# Exercises

There has been a lot of conversation recently about the impact of graduating college, i.e., obtaining a bachelor's degree, on one's future career and lifetime earnings. The common sentiment is that individuals who have earned a bachelor's degree (or higher) will earn more income over the course of a lifetime than an individual who does not have such a degree.

The culmination of individual factors can impact the wealth and resources of a county. Therefore we will examine the association between these two factors at a county level, using the percent of adults 25 years old and older with a Bachelor's degree to understand variability in median income.

::: callout-important
All narrative should be written in complete sentences, and all visualizations should have informative titles and axis labels.
:::

::: callout-tip
Write all code and narrative in your Quarto file. Throughout the assignment, you should periodically **render** your Quarto document to produce the updated HTML file.

Make sure we can read all of your code in your HTML file. This means you will need to break up long lines of code. One way to help avoid long lines of code is to start a new line after every pipe (`|>`).

In your `hw-04.qmd` document you'll see that each task is wrapped in an `exercise` box, with an empty `answer` box below it for you to fill in your response. Use this format for the remaining exercises.
:::

## Part 1: Exploratory data analysis

::: {.exercise}
Load `data/us-counties-sample.csv` into a data frame called `county_data`. Use `glimpse()` to take a look.

How many counties are in the data set?
:::

::: {.answer}

```{r read-data}
#| message: false

```

:::

::: {.exercise}
Create a histogram of the distribution of `bachelors` and calculate appropriate summary statistics with `favstats()`. Use the visualization and summary statistics to describe the distribution. Include an informative title and axis labels on the plot.
:::

::: {.answer}

```{r hist-bachelors}

```

:::

::: {.exercise}
Create a visualization of the relationship between `bachelors` and `median_household_income`, and use the `cor()` function to calculate the correlation. Use the visualization and correlation to describe the relationship between the two variables.
:::

::: {.answer}

```{r scatter-bachelors}

```

:::

::: callout-tip
This is a good place to render your document.
:::

## Part 2: Fitting the model

::: {.exercise}
Write the form of the statistical model using the template below. Use mathematical notation (i.e., with Greek letters) and variable names (`bachelors` and `median_household_income`) in the equation. I'm not asking you to fit the model yet — replace each word in the equation below with either the name of a variable or a Greek letter.

$$Response = intercept + slope\times predictor + error$$

::: callout-tip
[Click here](https://oeis.org/wiki/List_of_LaTeX_mathematical_symbols) for a guide on writing mathematical symbols using LaTeX. You will need to use a backslash (\\) before each underscore in the LaTeX code. For example, `avg_air_temp` will be written as *avg\\\_air\\\_temp.*
:::
:::

::: {.answer}

:::

::: {.exercise}
Fit the regression line corresponding to the statistical model in the previous exercise, and call it `income_mod`. Use `tidy()` to neatly display the model output using 3 digits.

Write the equation of the fitted model using mathematical notation. Use variable names (`bachelors` and `median_household_income`) in the equation.
:::

::: {.answer}

```{r fit-income-mod}

```

:::

::: {.exercise}
-   Interpret the slope of `income_mod`. The interpretation should be written in a way that is meaningful in the context of the data.
-   Is it useful to interpret the intercept for this data? If so, write the interpretation in the context of the data. Otherwise, briefly explain why not.
:::

::: {.answer}

:::

::: callout-tip
This is a good place to render your document.
:::

## Part 3: Residuals

::: {.exercise}
Choose one county from `county_data` (use `filter()` with its `name`) and record its actual `bachelors` and `median_household_income` values.

-   Using the fitted equation from Exercise 5, compute this county's **predicted** median household income by hand (i.e., by plugging into the equation, not with a function).
-   Compute the **residual**: the difference between the actual and predicted income.
-   Is this county above or below the regression line? Did the model over- or underpredict this county's income?
:::

::: {.answer}

```{r one-county}

```

:::

::: {.exercise}
Use `augment()` on `income_mod` to get fitted values and residuals for every county, and save the result to `income_aug`.

-   Use `arrange()` to find the county with the **largest positive** residual and the county with the **largest negative** residual.
-   Interpret each of these two residuals in the context of the data (what does it mean for that county's actual vs. predicted income?).
:::

::: {.answer}

```{r augment-income}

```

:::

::: callout-tip
This is a good place to render your document.
:::

## Part 4: Variation, $R^2$, and RMSE

::: {.exercise}
Fit the **empty model**, `empty_mod <- lm(median_household_income ~ NULL, data = county_data)`.

-   Using `augment()` and the residuals, compute the $SST$ (the SSE of the empty model) and the $SSE$ of `income_mod`.
-   Use subtraction to find the $SSM$ for `income_mod`, and then compute $R^2 = SSM/SST$ by hand.
-   Check your work using `supernova(income_mod)`.
:::

::: {.answer}

```{r ss-decomposition}

```

:::

::: {.exercise}
Use the `rsq()` and `rmse()` functions from `yardstick` on `income_aug` to compute $R^2$ and RMSE for `income_mod`.

-   Confirm that this $R^2$ matches what you computed by hand in Exercise 9, and that it equals the correlation from Exercise 3, squared.
-   Write one sentence interpreting $R^2$ and one sentence interpreting RMSE, in the context of the data.
:::

::: {.answer}

```{r rsq-rmse-income}

```

:::

::: {.exercise}
Fit a second model, `computer_mod`, predicting `median_household_income` from `household_has_computer`. Use `augment()`, then `rsq()` and `rmse()` to evaluate it.

Fill in the table below comparing the two models.

| Model | Predictor | $R^2$ | RMSE |
|:---|:---|:---:|:---:|
| `income_mod` | `bachelors` |  |  |
| `computer_mod` | `household_has_computer` |  |  |

Which model does a better job predicting median household income? Explain how you decided, using both $R^2$ and RMSE.
:::

::: {.answer}

```{r fit-computer-mod}

```

:::

::: callout-tip
This is a good place to render your document.
:::

::: {.exercise}
Create a new column in `county_data` called `income_thousands`, equal to `median_household_income` divided by 1000. Refit `income_mod` using `income_thousands` as the response instead, and recompute its $R^2$ and RMSE.

-   Which of the two numbers changed, and which stayed the same?
-   Explain why, in terms of what $R^2$ and RMSE each measure and what units they're in.
:::

::: {.answer}

```{r income-thousands}

```

:::

::: callout-tip
You're done and ready to submit your work! Render your document and upload the .qmd and .html files to Canvas.
:::

# Grading (30 pts)

<br>

| Component | Points |
|:---|:---|
| Ex 1 | 2 |
| Ex 2 | 2 |
| Ex 3 | 2 |
| Ex 4 | 2 |
| Ex 5 | 2 |
| Ex 6 | 2 |
| Ex 7 | 3 |
| Ex 8 | 3 |
| Ex 9 | 3 |
| Ex 10 | 3 |
| Ex 11 | 2 |
| Ex 12 | 2 |
| Grammar & Writing | 1[^1] |
| Workflow & formatting | 1[^2] |

[^1]: The "Grammar & Writing" grade is decided based on your grammar and writing. This is typically decided by choosing one of the questions and assessing the writing.

[^2]: The "Workflow & formatting" grade is to assess the reproducible workflow and document format. This includes having a neatly organized document with readable code and your name and the date in the YAML.
