---
title: "HW 03: The Coffee Truck"
author: "Insert Name Here"
subtitle: "Regression with quantitative and categorical predictors"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        </style>
---

```{r}
#| include: FALSE
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

# Introduction

This HW builds on the Coffee Truck activity from lecture: instead of playing with just two levels of one setting, you'll design and play a bigger game that lets you fit two different regression models, one with a quantitative predictor and one with a categorical predictor.

::: callout-note
R is the name of the programming language itself and RStudio is a convenient interface.
:::

## Learning goals

By the end of this homework, you will...

- Design and generate your own data set from the Coffee Truck game
- Clean a variable that R read in as text when it should be a number
- Fit, interpret, and use a simple linear regression model with a quantitative predictor
- Fit, interpret, and use a simple linear regression model with a categorical predictor, including choosing and changing the reference level
- Compare what each type of model can and cannot tell you

# Getting Started

### Log in to RStudio

- Go to [RStudio](https://rstudio.collegeofidaho.edu/) and login with your College of Idaho Email and Password.

- Make a folder called `hw` to store all of your homework for the semester, if you haven't already. Create a subfolder called `hw03` to store this homework, and a `data` subfolder inside of it.

- Log into [Canvas](<>), navigate to HW-03, and download the `hw-03.qmd` file. Upload it into the folder you just made. You will generate your own data set below and upload it into the `data` folder.

## R and RStudio

See the [R resources](https://ericfriedlander.github.io/math2025-fa26/computing-r-resources.html) page for a guide to the RStudio IDE and Quarto file components.

### YAML

The top portion of your Quarto file (between the three dashed lines) is called **YAML**. It stands for "Yet Another Markup Language". It is a human friendly data serialization standard for all programming languages. All you need to know is that this area is called the YAML (we will refer to it as such) and that it contains meta information about your document.

::: callout-important
Open the Quarto (.qmd) file in your project, change the author name to your name, and render the document. Examine the rendered document. You can now work off of that file.
:::

# Packages

We will use the following packages in today's HW.

```{r load-packages}
#| message: false
library(tidyverse)
library(mosaic)
library(ggformula)
library(broom)
library(coursekata)
```

# Data

You'll generate your own data by playing the [Coffee Truck game](https://www.stat2games.sites.grinnell.edu/games/coffeetruck.html) from Grinnell's **stat2games** project, the same simulator from the categorical predictors lecture and activity.

> Jo inherited a coffee truck, has \$20,000 left after repairs, and wants help figuring out what combination of settings sells the most coffee. Every simulated shift you play records:

- `Sales`: number of cups sold that shift (our response variable)
- `Location`: Business District / City Park / City Hall / Near the Zoo
- `Music`: No Music / Hip Hop / Alternative
- `Price`: \$2 / \$3 / \$4 / \$5 per cup
- `TimeOfDay`: Morning / Lunch / Afternoon / Evening

The CSV the game exports also includes a few columns we won't use in this HW, like `GrossIncome`, `Costs`, and `Profit`. You can ignore those.

::: callout-important
This game's data lands in a table that's **public to anyone**. Do not use your real name as your Player Name.
:::

## Analysis goal

In lecture, we fit a model using a categorical predictor while holding `Price` fixed. But `Price` is itself something Jo can set, and it's a number, not a category. In this HW you'll design a game that lets you look at `Price` as a quantitative predictor, and lets you look at one other setting (`Music` or `Location`) as a categorical predictor with _all_ of its levels, then fit one model of each type.

# Exercises

::: callout-tip
Write all code and narrative in your Quarto file. Write all narrative in complete sentences. Throughout the assignment, you should periodically **render** your Quarto document to produce the updated HTML file.

Make sure we can read all of your code in your HTML file. This means you will need to break up long lines of code. One way to help avoid long lines of code is to start a new line after every pipe (`|>`).

In your `hw-03.qmd` document you'll see that each task is wrapped in an `exercise` box, with an empty `answer` box below it for you to fill in your response. Use this format for the remaining exercises.
:::

::: {.exercise}
Play the game at the link above.

- On the ID screen, enter **`MATH2025FA26`** as the Group Name.
- Pick a Player Name that does **not** identify you, and write it in the box below (you'll need it later).

Now set up your design:

- Set **Price** to select **all four levels** (\$2, \$3, \$4, \$5).
- Pick **one** of **Music** or **Location**, and select **all of its levels** (3 for Music, 4 for Location). Leave the other one at a single, fixed level.
- Hold **Time = Lunch**.
- Set **Design Type** to `Random Sample` and **Number of Repetitions** to the maximum allowable (`3` or `5` depending on which variable you selected).
- Click **Start Sim**.

Record your Player Name, which variable you chose (Music or Location), the levels you selected, and the number of days from your design.
:::

::: {.answer}

:::

::: {.exercise}
Download your data as a CSV file, and upload it into the `data`. I recommend changing the filename to something more memorable. Then use `read_csv` to load it and take a look.

- How many rows and columns does it have? Does the row count match the design you recorded in Exercise 1?

:::
::: {.answer}

```{r read-data}
#| message: false

```

:::

::: {.exercise}
You should notice that `Price` was read in as a `chr` (character) column instead of `dbl` (double), the data type for numeric data.

- Why did `Price` get read in as a character data type instead of a double? Be specific.
- Why do we need `Price` to be numeric before we can use it as a quantitative predictor?

Write some code below that fixes this. Hint: It should be the same way we fixed `spend_per_resident_data` in HW 02.
:::

::: {.answer}

```{r clean-price}

```

:::

::: callout-tip
This is a good place to render your document.
:::

# Model 1: Price as a quantitative predictor

::: {.exercise}
Create a plot of `Sales` axis against `Price` with a fitted line added via `gf_lm()`. Give the plot an informative title and axis labels.

Describe the relationship you see between `Price` and `Sales`.
:::

::: {.answer}

```{r plot-price}

```

:::

::: {.exercise}
Fit a linear model called `price_mod` predicting `Sales` from `Price`. Use `tidy()` to format the output, and write out the estimated regression equation.
:::

::: {.answer}

```{r fit-price-mod}

```

:::

::: {.exercise}

- Interpret the slope of `price_mod` in the context of the data.
- Interpret the intercept of `price_mod` in the context of the data. Is this interpretation useful in practice? Briefly explain, using the range of prices Jo can actually charge.

:::

::: {.answer}

:::

::: {.exercise}
Use `predict()` to get `price_mod`'s predicted `Sales` at prices of \$2.50, \$3.50, and \$4.50, prices the truck never actually charged.

```{r predict-price}
#| eval: false
new_prices <- tibble(Price = c(2.5, 3.5, 4.5))

## more code here
```

:::

::: {.answer}

:::

::: callout-tip
This is a good place to render your document.
:::

# Model 2: your categorical predictor

::: {.exercise}
Find the mean `Sales` at each level of your chosen variable (Music or Location). Then create a plot exploring the relationship between your variable and `Sales`, similar to the one from lecture.

Which level has the highest average `Sales`?
:::

::: {.answer}

```{r favstats-cat}

```

```{r plot-cat}

```

:::

::: {.exercise}
Fit a linear model called `cat_mod` predicting `Sales` from your variable. Use `tidy()` to format the output.

- What is the reference level in your model? How can you tell from the output alone?
- Write out the estimated regression equation using R's coefficient names (e.g. `MusicHipHop`), and give the dummy coding (the 0/1 values for each indicator) for an observation at each level of your variable by modifying the table below. You can find documentation on how make tables in Quarto [here](https://quarto.org/docs/authoring/tables.html)
  :::

::: {.answer}

```{r fit-cat-mod}

```

| Original variable | Col1    | Col2    | Col3     |
| ----------------- | ------- | ------- | -------- |
| Value1            | Value2  | Value3  | Value 4  |
| Value5            | Value6  | Value7  | Value 8  |
| Value9            | Value10 | Value11 | Value 12 |
| Value13           | Value14 | Value15 | Value 16 |

:::

::: {.exercise}
Interpret the intercept and each coefficient from `cat_mod` in the context of the data.
:::

::: {.answer}

:::

::: {.exercise}
Compare the group means from Exercise 8 to the coefficients from `cat_mod` (Exercise 9). Show, using the coefficients, how to recover the predicted `Sales` for each level of your variable, and confirm it matches the corresponding mean.

How many distinct predictions can `cat_mod` make, and why?
:::

::: {.answer}

:::

::: callout-tip
This is a good place to render your document.
:::

# Putting it together

::: {.exercise}
Suppose, instead of treating `Price` as quantitative in Model 1, you had treated it as a categorical variable with four levels (\$2, \$3, \$4, \$5).

- How many coefficients would that model have (including the intercept)?
- What could that model do that `price_mod` cannot?
- What could `price_mod` do that a categorical-price model cannot?

No code needed for this question.
:::

::: {.answer}

:::

::: {.exercise}
Using `price_mod` and `cat_mod` separately (you have not been asked to combine them into one model):

- What price would you recommend to Jo, based on `price_mod`?
- What level of your variable would you recommend to Jo, based on `cat_mod`?
- What can these two _separate_ models **not** tell you about that price/level combination together?
  :::

::: {.answer}

:::

::: {.exercise}

- `Sales` is the number of cups sold, not profit. Why might the price that maximizes `Sales` not be the price that maximizes Jo's profit?
- Why might a linear model not be the best model to choose when trying to predict `Profit` from price? Hint: What happens to the number of Sales when the price changes?

:::

:::{.answer}

:::

::: callout-tip
You're done and ready to submit your work! Render your document and upload the .qmd and .html files to Canvas.
:::

# Grading (30 pts)

<br>

| Component             | Points |
| :-------------------- | :----- |
| Ex 1                  | 1      |
| Ex 2                  | 2      |
| Ex 3                  | 2      |
| Ex 4                  | 2      |
| Ex 5                  | 2      |
| Ex 6                  | 2      |
| Ex 7                  | 2      |
| Ex 8                  | 2      |
| Ex 9                  | 3      |
| Ex 10                 | 2      |
| Ex 11                 | 2      |
| Ex 12                 | 2      |
| Ex 13                 | 2      |
| Ex 14                 | 2      |
| Grammar & Writing     | 1[^1]  |
| Workflow & formatting | 1[^2]  |

[^1]: The "Grammar & Writing" grade is decided based on your grammar and writing. This is typically decided by choosing one of the questions and assessing the writing.

[^2]: The "Workflow & formatting" grade is to assess the reproducible workflow and document format. This includes having a neatly organized document with readable code and your name and the date in the YAML.
