---
title: "AE 02: Multivariate Relationships"
author: "Your name here"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        </style>
---

::: callout-important
-   For this exercise, you will work in groups, but everyone will work through the document and submit their solution.

-   Open **RStudio** and create a subfolder in your AE folder called "AE-02"

-   Go to the [Canvas](insert link) and locate your `AE 02` assignment to get started.

-   Upload the `ae-02.qmd` files into the folder you just created. To submit this Activity, upload the `.qmd` and `.html` responses in the corresponding Canvas assignment.
:::

```{r}
#| include: FALSE
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

# Multivariate Relationships and Exploratory Data Analysis (EDA) with R

## Carbohydrates and Starbucks

Starbucks often displays the total calories in their food items but not the other nutritional information. Carbohydrates are a body's main fuel source. The [Dietary Guidelines for America](https://www.mayoclinic.org/healthy-lifestyle/nutrition-and-healthy-eating/in-depth/carbohydrates/art-20045705) recommend that carbohydrates make up 45% to 65% of total daily calories. Our goal is to understand the relationship between the amount of carbohydrates and calories in Starbucks food items. We'd also like to assess if the relationship differs based on the type of food item (bakery, salad, sandwich, etc.)

You will now be introduced to the functions we will use most frequently for exploratory data analysis (EDA) in this class.
The first part of this document contains explanations for beginning R users, but **all students should work through the entire document**.


::: callout-important
Reminder: I suggest "rendering as you go": rendering every few chunks, to make sure things are rendering correctly, rather than waiting until the end of a document to render the whole thing (and potentially encountering lots of errors that you have to unpack)! I also ALWAYS suggest "saving as you go": it's a good idea to save every couple of minutes. (This is good practice for all your files.)
:::

## Loading Packages

R is an **open source** programming language and environment for statistical computing and graphics.
This means that anyone can use R for free, and anyone can contribute to the development of R.
R has a large ecosystem of packages that extend its functionality.
You can think of these like expansion packs for R.
If you're on your own computer, each packages needs to be install once using the function `install.packages()`.
Since we're working in the cloud, we don't need to do this step.
Regardless of what computer you're on each package needs to be loaded using the function `library()` every timne you want to use it
We typically load all packages at the beginning of our scripts and Quarto files.

::: {.exercise}

The code below loads the package called `tidyverse` which provides many functions that will be useful for manipulating data and the package called `mosaic` which is useful for teaching statistics. Add another line that loads a package called `ggformula` which we will use to plot our data and a third that loads the package `openintro` which will contain our data for this activity. Note that `#| warning: FALSE` causes the code to run without displaying any associated warnings when you render. You will often see `#| eval: FALSE` which prevents the code chunk from running when you render the document. Throughout this course, make sure you remove these lines or change them to `#| eval: TRUE` whenever you reach them. The reason that are included in the first place is so I can write scaffolded and fill-in-the-blank type exercises for you. These will cause errors (since they are incomplete code) until you finish writing them.
:::

::: {.answer}
```{r}
#| warning: FALSE
#| eval: FALSE

# load packages we typically use for this class.
library(tidyverse)
library(mosaic)
library(_________) # load ggformula
# load openintro
```
:::

## Loading data

Doing statistics requires *data*. Sometimes we will have to load data from a file and sometimes we can get our data from an R package. Today's dataset, called `starbucks` is from the `openintro` package. To load a this data from within R you can use the `data` command.

::: {.exercise}
Use the code below to load the dataset `starbucks`, which is part of the `openintro` package. Don't forget to change `eval: FALSE` to `eval: TRUE`... Why do you think we need `eval: FALSE` if you aren't actually rewriting any code?
:::

::: {.answer}
```{r}
#| eval: FALSE

data("starbucks")
```
:::

This is a great chance to remind you that R cares about letter case. This means that `data("starbucks")` works but `data("Starbucks")` doesn't!

### Inspecting the data

Now you're ready to learn a little bit about the `starbucks` data set.

::: {.exercise}
Run the code below and then edit the bulleted list below to add a short description in your own words describing what each function does. Hint: if it isn't clear based on the output you can use `?` before the function name in the console to bring up the documentation on that function.
:::

::: {.answer}
```{r}
#| eval: FALSE

# Inspecting the data
glimpse(starbucks)
head(starbucks)
names(starbucks)
```

-   `glimpse()` : this function...
-   `head()` : this function...
-   `names()` : this function...
:::

::: {.exercise}
Based on the output above, how many **observations** and how many **variables** does this data set have?
:::

::: {.answer}

:::

::: {.exercise}
Let's do a little bit of a data prep. In the **console** type `?starbucks` and look at the definition of `type`.

The following is a little bit of data wrangling to get the source data in shape for our purposes. What do you think this code is doing? Make sure you change `eval: FALSE` to `eval: TRUE`.

```{r}
#| eval: FALSE

starbucks <- starbucks |>
  mutate(bakery = factor(if_else(type == "bakery", "bakery", "non-bakery")))
```

Statisticians should always know something about the data domain in order to be useful. Why do you think we are changing our data like this?
:::

::: {.answer}

:::

## Exploratory Data Analysis

For the purposes of our class, it's useful to learn a model-centric approach to R. The pseudo-code below is going to be our foundation for the rest of the class:

`function( Y ~ X, data = DataSetName )`

Here's a short description of each part in the pseudo-code above:

-   `function` is an R function that dictates something you want to do with your data; for example,
    -   `mean` calculates the mean
    -   `lm` fits a linear regression model
-   `Y` is the outcome of interest (response variable)
-   `X` is some explanatory variable; you can use `1` as a placeholder if there is no explanatory variable
-   `DataSetName` is the name of a data set loaded into the R environment

Always start with clear research questions. Our question for this exercise:

> *How do the number of carbs compare for `bakery` and `non-bakery` items?*

The purpose of the exploratory data analysis (EDA) is to learn as much as you can about your research question **before doing any statistical modeling**. We basically want to try and answer the research question with EDA if possible...or at least have a guess as to what the answer "should" be. Then we use statistical models to formally accommodate variability in the data and calculate the uncertainty of our conclusions.

::: {.exercise}
Use the R code chunk below to calculate the mean carbs by for bakery and non-bakery items. Summarize your observations below the code chunk.

```{r}
#| eval: FALSE

mean(carb ~ bakery, data = starbucks)
```
:::

::: {.answer}

:::

::: {.exercise}
Of course, there are lots of other ways to summarize a numerical variable besides the mean. Create an R code chunk below and use the function `favstats` to calculate the other summary statistics for the number of carbs for `bakery` and `non-bakery` items.

Summarize your observations below the code chunk.

(Hint: it should look a lot like the code chunk above with `mean` replaced by `favstats`.)
:::

::: {.answer}

:::

### Univariate plots

Let's generate some univariate (i.e. one-variable) EDA plots for our three variables. In general, can use histograms or boxplots (`gf_histogram` or `gf_boxplot`) for quantitative variables and bar charts `gf_bar` for categorical variables.

::: {.exercise}
Fill in the blanks in the code below to generate the corresponding plots. Then create a boxplot of `carb`. Finally, share your observations below the code:
:::

::: {.answer}
```{r}
#| eval: FALSE

# make a histogram of carb
gf_histogram(~ carb, data = starbucks)

# make a histogram of calories
gf_histgoram(~ _____, data = ______)

# make a bar chart of bakery
gf_bar(~ _____, data = _____)

# make a boxplot of carb
```

:::

### Bivariate plots

::: {.exercise}
Let's now generate some bivariate (i.e. two-variable) EDA plots for our three variables. For two-quantitative variables we typically use a scatter plot `gf_point`. If we are comparing a quantitative and a categorical variable will use side-by-side boxplots `gf_boxplot`.
Generate and plot to visualize `carb` and `calories` and one to visualize `carb` and `bakery`. Comment on your observations.
:::

::: {.answer}

:::

## Submitting Your Assignment

:::{.exercise}
1. Click **Render** one last time to render your document.
2. Review the html output to make sure it reflects what you want it to say.
3. Upload the `qmd` and `html` files to Canvas.
:::
