---
title: "AE 03: Bike rentals in Washington, DC"
subtitle: "Simple linear regression"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        </style>
---

::: callout-important
-   Open RStudio and create a subfolder in your AE folder called "AE-03"

-   Go to the [Canvas]() and locate the `AE-03` assignment. That link should take you directly there.

-   Upload the `ae-03.qmd` into the folder you just created
-   Upload `dcbikeshare.csv` into your `data` folder
:::


:::{.exercise}
Run the following code to load the packages we'll use in this activity.
:::

:::{.answer}
```{r load-packages}
#| message: false
library(tidyverse)
library(broom)
library(ggformula)
library(patchwork)
```
:::

## Data

Our data set contains daily rentals from the Capital Bikeshare in Washington, DC in 2011 and 2012. It was obtained from the `dcbikeshare` data set in the `dsbox` R package.

We will focus on the following variables in the analysis:

-   `count`: total bike rentals
-   `temp_orig`: Temperature in degrees Celsius
-   `season`: 1 - winter, 2 - spring, 3 - summer, 4 - fall

Click [here](https://tidyverse.github.io/dsbox/reference/dcbikeshare.html) for the full list of variables and definitions.

## Exercises

::: {.exercise}
Modify the file path to load the `dcbikeshare` data set.

Hints:

- Use a relative path
- You need to go up two levels and then into the `data` folder
- Pressing TAB while your cursor is in the quotations will help you navigate
- Make sure to remove the `#|eval: false` line from the answer code
:::

::: {.answer}
```{r load-data}
#| message: false
#| eval: false
bikeshare <- read_csv("insert path here")
```
:::

::: {.exercise}
Below is code to create visualizations of the distributions of daily bike rentals and temperature.
Create a plot of the number of rentals vs the temperature and store it in a variable called `p3`.
If you can figure it out based on context, try and add appropriate labels to the plot.
Finally, uncomment the last line of code in this chunk. What do you think this last line does? Comment on what you see.
:::

::: {.answer}
```{r bike-temp-viz}
#| warning: false
#| message: false
#| eval: false

p1 <- gf_histogram(~ count, data = bikeshare, binwidth = 250) |>
  gf_labs(x = "Daily bike rentals")

p2 <- gf_histogram(~temp_orig, data = bikeshare) |>
  gf_labs(x = "Temperature (Celsius)")

# (p1 | p2) / p3
```
:::

::: {.exercise}
In the raw data, seasons are coded as 1, 2, 3, 4 as numerical values, corresponding to winter, spring, summer, and fall respectively. Recode the `season` variable to make it a categorical variable (a factor) with levels corresponding to season names, making sure that the levels appear in a reasonable order in the variable (i.e., not alphabetical).
:::

::: {.answer}
```{r recode-season}
# add code developed during livecoding here
```
:::

::: {.exercise}
We want to evaluate whether the relationship between temperature and daily bike rentals is the same for each season. To answer this question, first create a scatter plot of daily bike rentals vs. temperature faceted by season.

```{r facet-plot}
# add code developed during livecoding here
```
:::

::: {.answer}

```{r}

```

:::

::: {.exercise}
-   Which season appears to have the **strongest** relationship between temperature and daily bike rentals? Why do you think the relationship is strongest in this season?

-   Which season appears to have the **weakest** relationship between temperature and daily bike rentals? Why do you think the relationship is weakest in this season?
:::

::: {.answer}

:::

::: {.exercise}
Going back to the scatter plot of rentals vs. temperature from earlier, there appears to be one day with a very small number of bike rentals. Use `filter` to figure out what that day was? Why were the number of bike rentals so low on that day? *Hint: You can Google the date to figure out what was going on that day.*
:::

::: {.answer}

:::

::: {.exercise}
Filter your data for the season you've been assigned. Give the resulting data set a new name.
:::

::: {.answer}
```{r filter-season}
# add code developed during livecoding here
```
:::


::: {.exercise}
Using the subset of the data from the previous exercise, fit a linear model for predicting daily bike rentals from temperature for this season.
:::

::: {.answer}
```{r}

```
:::

::: {.exercise}
Use the output to write out the estimated regression equation on the whiteboard.
:::

::: {.exercise}
Interpret the slope in the context of the data.
:::

::: {.answer}

:::

::: {.exercise}
Interpret the intercept in the context of the data.
:::

::: {.answer}

:::

::: {.exercise}
Create a visualization of the model we just created.
:::

::: {.answer}
```{r}

```
:::

::: callout-important
To submit the AE:

-   Render the document to produce the HTML with all of your work from today's class.
-   The driver for your group should upload your .qmd and .html files to the Canvas assignment.
:::
