---
title: "AE 07: Life Expectancy on the Islands"
subtitle: "Sampling a population and modeling with a categorical predictor"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

<!-- TODO (instructor): fill in the Canvas link below and the published Google Sheet CSV link in Exercise 3 before class. -->

::: callout-important
- Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-07"
- Go to the assignment for this activity in [Canvas]() to get started.
- Upload the `ae-07-islands-life-expectancy.qmd` file into the folder you just created. We will collect the data for this activity together.
:::

## Intro

What is the life expectancy on the Island? Is it different for men and women?

We can't ask every islander, so we'll do what a previous class did: take a **stratified cluster sample** of the Island's records, then **fit a model predicting life expectancy from sex**. Each group is responsible for a **cluster** of villages, and everyone's data goes into one shared Google Sheet.

Here is the sampling plan for **today**. This is a simplified version of the previous class's approach. They built age-specific mortality rates (a life table) to compute life expectancy, but today we'll use the **mean age at death among islanders who have died** as a stand-in for life expectancy.

::: callout-important
Your data goes into a sheet that everyone in the class can see and edit. **Only add rows, never delete or overwrite anyone else's.** Enter `sex` as exactly `Female` or `Male` (capitalized, no spaces) so the pooled data stays clean.
:::

## Exercises

### Getting set up and collecting data

::: exercise
Log into [The Islands](https://islands.smp.uq.edu.au/index.php) with your College of Idaho email. Within your group, assign the roles below, and record who has which role.

- **Navigator:** finds islanders on the Islands
- **Recorder:** enters rows into the shared Google Sheet
- **Coder:** runs the R code in this document

Find your group's villages in the table below and locate them on the Islands.

| Group | Villages |
|:--|:--|
| A | Fairhaven, Serrano, Sunrise, Eden |
| B | Riverside, Valkov, Rosica |
| C | Edwardton, Asderaki, Mossgiel |
| D | Macondo, Durand Bay, Deep Bay |

Before you start collecting: what **type of sampling** is this (simple random, stratified, cluster, or a combination)? Which part of the plan is the **cluster**?
:::

::: answer

:::

::: exercise
Now collect your data. Within **each of your group's villages**, sample islanders who **have died**. Record one row per islander in the shared Google Sheet with these columns:

| `village` | `sex` | `birth_year` | `death_year` | `recorder` |
|:--|:--|:--|:--|:--|

Some rules, so that we all collect data the same way:

- Only include islanders who died in **1980 or later**, and skip anyone whose age at death would be over **150**.
- Try to spread your samples across all of your group's villages, rather than taking everyone from the first one you find.
- Choose islanders **randomly** (for example, by picking a random spot in the list rather than the first few).
- Aim for about **25 islanders** per group. Remember that each islander can only do one task at a time.

Why do you think we restrict to islanders who died recently (1980 or later)? Why might it matter *how* we choose which islanders to record?
:::

::: answer

:::

### Loading and cleaning the pooled data

::: exercise
Load the packages below, then read the pooled data into R using `read_csv()` and the published link to the shared Google Sheet:

**[INSTRUCTOR: paste the published Google Sheet CSV link here]**

a.  Use `glimpse()` to look at your data. How many islanders are in the pooled sample?

b.  Create a new column called `age_at_death` equal to `death_year - birth_year`. Then keep only the rows that follow our rules: `death_year` of 1980 or later, and `age_at_death` between 0 and 150. Save the result as `islands`.

c.  How many rows did the cleaning step remove? Look at a few of them. Why might they have been removed?
:::

::: answer
```{r load-packages}
#| message: false
library(tidyverse)
library(mosaic)
library(ggformula)
library(broom)
library(coursekata)
```

```{r load-data}
#| message: false

```

```{r clean-data}

```
:::

### Estimating life expectancy

::: exercise
Explore `age_at_death`.

a.  Use `favstats()` to compute summary statistics for `age_at_death` overall, and then separately for each level of `sex`.

b.  Make a plot exploring the relationship between `sex` and `age_at_death` (a boxplot works well).

c.  In a sentence or two, report the **overall mean age at death**. This is our estimate of life expectancy on the Island. Then describe what the plot and the group summaries say about differences between men and women.
:::

::: answer
```{r eda}

```
:::

::: exercise
Fit a linear model called `le_fit` that predicts `age_at_death` from `sex`. Use `tidy()` to display the output.

- What is the reference level? How can you tell from the output alone?
- Write out the estimated regression equation.
:::

::: answer
```{r fit-model}

```
:::

::: exercise
Interpret the intercept and the slope of `le_fit` in the context of the data.

Then confirm, using your group summaries from the earlier exercise, that the slope equals the **difference in the group means**.
:::

::: answer
```{r check-slope}

```
:::

### Thinking critically

::: exercise
Let's think about what this sample can tell us.

- What **population** does our sample actually represent? Is it the whole Island? All islanders who have ever lived? Something else?
- Name **two ways** our estimate of life expectancy could be biased. (Think about who is *in* our sample, and who is missing.) One hint: think about who cannot possibly appear among "islanders who have died."
- Was this an **observational study** or an **experiment**? What would we be able to conclude from `le_fit`, and what *couldn't* we conclude?
- Could you design an *experiment* on the Islands to find out whether sex affects life expectancy? Would you want to?
:::

::: answer

:::

::: callout-note
**Stop here.** We'll come back together and talk about what our results say (and don't say) before you continue.
:::

### Is the difference real?

::: exercise
Each group will now refit the model using **only the rows from your own group's villages**.

a.  Use `filter()` to keep only rows where `village` is one of your group's villages, and save the result as `my_group`. Use the table in Exercise 1 if you need to check the village names.

b.  Fit `lm(age_at_death ~ sex, data = my_group)` and use `tidy()` to display it.

c.  Write your group's slope on the whiteboard next to the slopes from the other groups (and next to the slope from the pooled data).
:::

::: answer
```{r group-fit}

```
:::

::: exercise
Look at the slopes on the whiteboard.

- How do they compare with each other? With the pooled slope?
- Given how much the numbers differ, how confident are you that men and women on the Island **really** have different life expectancies?
- What could you do to be *more* confident? What would you need to know about how much these slopes typically vary?
:::

::: answer

:::

::: callout-important
To submit the AE:

- Render the document to produce the HTML with all of your work from today's class.
- The driver for your group should upload your .qmd and .html files to the Canvas assignment.
:::
