---
title: "AE 09: Collecting the Islands Data"
subtitle: "Carrying out a stratified cluster sample"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

::: callout-important
- Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-09".
- Go to the assignment for this activity in [Canvas](https://cofi.instructure.com/courses/18743/assignments/244041) to get started.
- Upload the `ae-09-islands-collection.qmd` file into the folder you just created.
:::

## Intro

What is the life expectancy on the Island? Is it different for men and women?

We can't ask every islander, so we'll do what a previous class did: take a **stratified cluster sample** of the Island's records. Today we collect the data, think about who ends up in it, and clean the pooled data together. Next class (AE 10), we explore the pooled data and ask whether sex explains **more** variation in age at death than we'd expect by chance.

Instead of computing life expectancy properly with a life table, we'll use the **mean age at death among islanders who have died** as a stand-in.

Here is the sampling plan:

1.  We stratify by **island**, so every island is represented.
2.  Within an island, the **villages** are the clusters. It would take too long to sample every village, so each of you is randomly assigned one.
3.  In your village, open the Town Hall, click **Records**, and open the **375-399** link. This lists every islander in your village who died between Island years 375 and 399.
4.  Randomly select **10** of those islanders and record their information in the shared spreadsheet.

::: callout-important
Your data goes into a spreadsheet that everyone in the class can see and edit. **Only add rows, never delete or overwrite anyone else's.** Enter `sex` as exactly `Female` or `Male`, and enter every value the same way each time.
:::

## Part 1: Collecting the data

### Getting oriented

::: exercise
Log into [The Islands](https://islands.smp.uq.edu.au/index.php) with your College of Idaho email. Find the village you were assigned, and record its **name** and **island** here.

Before you start collecting: what **type of sampling** is this (simple random, stratified, cluster, or a combination)? Which part of the plan is the **stratification**, and which part is the **cluster**?
:::

::: answer
:::

### Collecting your sample

::: exercise
Open the **Records** for your village and click the **375-399** link. Then open [the shared Excel sheet](https://yotescollegeofidaho-my.sharepoint.com/:x:/g/personal/efriedlander_collegeofidaho_edu/IQB2Q0ijZYPzQYuumjlk_x09AQaTTMpIIkDa73TeNCMKJiI?e=16bSqH), create a new worksheet (the + button at the bottom), rename it with your village's name, and copy and paste the list of islanders into it.

How many islanders are in your village's list? (Look at the number of rows.)
:::

::: answer
:::

::: exercise
Randomly select **10** islanders from your list by running the code below. Fill in your birthday in `MMDDYYYY` format (for example, `12181988`) and the number of islanders in your list.

The numbers it prints are the **row numbers** of the islanders you will sample.

```{r}
#| eval: false

birthday <- # put your birthday here in MMDDYYYY format (e.g. 12181988)
set.seed(birthday)
number_of_villagers <- # put your total number of islanders here
sample(1:number_of_villagers, 10)
```

Why do we use `set.seed()` here? Is choosing islanders this way a *random* sample?
:::

::: answer
```{r draw-sample}
#| eval: false

```
:::

::: exercise
Use the row numbers to look up your 10 islanders. In the **MASTER DATASET** worksheet of the same shared Excel sheet, add one row per islander with:

- your name
- the island
- the village
- the islander's name
- sex
- birth year
- death year

Remember: only add rows.

Note: Your rows do not need to be adjacent to one another.
:::

::: answer
:::

### Thinking about who is in the sample

::: exercise
We are only sampling islanders who died between Island years 375 and 399.

- Who **cannot** possibly appear in this sample?
- Why might this make our data a poor stand-in for the life expectancy of a *newborn* today?
- Name one way our sampling plan is better than just picking the first 10 islanders we come across.

Time permitting, also think about (and be prepared to discuss):

1.  We are selecting the same number of villages from each Island and the same number of individuals from each village... could this bias our results?
2.  We aren't stratifying by sex. Perhaps it would've been a good idea to do so... Can you think of a situation in which this isn't a good idea? Hint: it may help if you imagine that the islanders are an alien species different than humans.
:::

::: answer
:::

::: callout-note
**Stop here.**

- We need to wait for you classmates to finished entering their data before we can move on.
- **If you finish early:** get a head start on [HW 05](https://ericfriedlander.github.io/math2025-fa26/hw/hw-05.html). It's about study design and doesn't need any data.
:::

## Part 2: Cleaning the pooled data

Let's clean the data together.

::: exercise
Load the pooled data as it stands right now.

a.  Use `glimpse()` and `count(sex)` to look at it. How many islanders are in it so far? Is each row one islander? Is `sex` spelled consistently?

b.  We'll clean the data together in class. Follow along and record the steps in the code chunk below, including creating `age_at_death` as `death_year - birth_year`. Save the cleaned result as `islands`.
:::

::: answer
```{r load-partial-data}
#| message: false
library(tidyverse)
library(mosaic)
library(ggformula)
library(broom)

islands_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTonqSnj8t3ZgjDzqCuq-P8bE-1OuHmwgBI0kF7eip7gkSmrJQGYRVwFtehtVS4FM6zwNAHUl1He29A/pub?output=csv")
```

```{r partial-clean}

```
:::

## Time permitting: Modeling differences by sex

Results will change as more rows come in. That's fine: we're previewing what we'll do next class.

::: exercise
Model the difference in age at death between men and women.

a.  Fit a linear model called `sex_fit` predicting `age_at_death` from `sex`, and display the results with `tidy()`.

b.  What is the reference level? Write out the estimated regression equation.

c.  Interpret the intercept and the slope in context.

d.  Compute the $R^2$ for the model and interpret it.
:::

::: answer
```{r sex-fit}

```
:::

::: callout-important
To submit the AE:

- Render the document to produce the HTML with all of your work from today's class.
- Upload your .qmd and .html files to the Canvas assignment.
:::