---
title: "AE 09: Analyzing the Islands Data"
subtitle: "Cleaning pooled data and modeling with a categorical predictor"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

<!-- TODO (instructor): fill in the Canvas link below, the link to the pooled data in Exercise 1, and adjust the cleaning steps to match how the class entered the data. -->

::: callout-important
- Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-09"
- Go to the assignment for this activity in [Canvas]() to get started.
- Upload the `ae-09-islands-analysis.qmd` file into the folder you just created.
:::

## Intro

In AE 08, everyone in the class sampled 10 islanders who died between Island years 375 and 399 from an assigned village. Today we pick up where we left off: we pool everyone's data, clean it, and answer our question.

What is the life expectancy on the Island? Is it different for men and women?

We'll use the **mean age at death among islanders who have died** as a stand-in for life expectancy, and we'll **fit a model predicting age at death from sex**.

## Exercises

### Loading and cleaning the pooled data

::: exercise
Load the packages below, then read the pooled data into R.

**[INSTRUCTOR: paste the link to (or instructions for getting) the pooled data here]**

a.  Use `glimpse()` to look at your data. How many islanders are in the pooled sample? Is each row one islander?

b.  Everyone typed in their own rows, so check for problems. For example, use `count(sex)` to see whether `sex` is spelled consistently, and look for missing values, duplicated rows, and columns stored as the wrong type. Fix what you find.

c.  Create a new column called `age_at_death` equal to `death_year - birth_year`. Then keep only the rows with a plausible `age_at_death` (between 0 and 150). Save the result as `islands`.

d.  How many rows did the cleaning step remove? Look at a few of them. Why might they have been removed?
:::

::: answer
```{r load-packages}
#| message: false
library(tidyverse)
library(mosaic)
library(ggformula)
library(broom)
library(coursekata)
```

```{r load-data}
#| message: false

```

```{r clean-data}

```
:::

### Estimating life expectancy

::: exercise
Explore `age_at_death`.

a.  Use `favstats()` to compute summary statistics for `age_at_death` overall, and then separately for each level of `sex`.

b.  Make a plot exploring the relationship between `sex` and `age_at_death` (a boxplot works well).

c.  In a sentence or two, report the **overall mean age at death**. This is our estimate of life expectancy on the Island. Then describe what the plot and the group summaries say about differences between men and women.
:::

::: answer
```{r eda}

```
:::

::: exercise
Fit a linear model called `le_fit` that predicts `age_at_death` from `sex`. Use `tidy()` to display the output.

- What is the reference level? How can you tell from the output alone?
- Write out the estimated regression equation.
:::

::: answer
```{r fit-model}

```
:::

::: exercise
Interpret the intercept and the slope of `le_fit` in the context of the data.

Then confirm, using your group summaries from the earlier exercise, that the slope equals the **difference in the group means**.
:::

::: answer
```{r check-slope}

```
:::

### Thinking critically

::: exercise
Let's think about what this sample can tell us.

- What **population** does our sample actually represent? Is it the whole Island? All islanders who have ever lived? Something else?
- Name **two ways** our estimate of life expectancy could be biased. (Think about who is *in* our sample, and who is missing.) One hint: think about who cannot possibly appear among "islanders who have died."
- Was this an **observational study** or an **experiment**? What would we be able to conclude from `le_fit`, and what *couldn't* we conclude?
- Could you design an *experiment* on the Islands to find out whether sex affects life expectancy? Would you want to?
:::

::: answer

:::

::: callout-note
**Stop here.** We'll come back together and talk about what our results say (and don't say) before you continue.
:::

### Does sex explain more variation than chance?

::: exercise
How much variation in age at death does `sex` explain in our data?

a.  Compute the PRE (also called $R^2$) for `le_fit` with `pre(age_at_death ~ sex, data = islands)`, and save it as `pre_obs`.

b.  Write a sentence interpreting `pre_obs` as a percent of the variation in age at death.

c.  Adding a predictor can never make PRE negative. Why not? Given that, why isn't "PRE is bigger than 0" convincing evidence that sex matters?
:::

::: answer
```{r pre-obs}

```
:::

::: exercise
Now let's imagine a world with **no relationship** between sex and age at death, just like the cards in class.

a.  Run `pre(shuffle(age_at_death) ~ sex, data = islands)` at least four times. Why does the answer change each time? Was it ever negative? Was it ever exactly 0?

b.  In your own words, what does `shuffle()` do to the relationship between `sex` and `age_at_death`? What does it *not* change?
:::

::: answer
```{r shuffle-once}

```
:::

::: exercise
Build the **null distribution** by shuffling 1000 times.

a.  Run `null_pre <- do(1000) * pre(shuffle(age_at_death) ~ sex, data = islands)`. Then make a histogram of the `pre` column of `null_pre` and add a vertical line at `pre_obs` (try `gf_histogram()` and `gf_vline()`).

b.  What does a single value in this histogram represent? What is the histogram a picture of?

c.  Compute the proportion of shuffles with a PRE at least as large as `pre_obs`: `mean(null_pre$pre >= pre_obs)`.

d.  Answer the question: does sex explain **more** variation in age at death than we'd expect if there were no relationship? Explain your reasoning, and say what your conclusion does and does *not* tell us (think back to how our sample was collected).
:::

::: answer
```{r null-dist}

```
:::

::: callout-important
To submit the AE:

- Render the document to produce the HTML with all of your work from today's class.
- Upload your .qmd and .html files to the Canvas assignment.
:::
