---
title: "AE 08: Collecting Data on the Islands"
subtitle: "Designing and carrying out a stratified cluster sample"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

<!-- TODO (instructor): fill in the Canvas link below before class. -->

::: callout-important
- Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-08"
- Go to the assignment for this activity in [Canvas]() to get started.
- Upload the `ae-08-islands-data-collection.qmd` file into the folder you just created.
- **Your credit for this AE is based on whether all 10 of your islanders are entered in the MASTER DATASET by the start of the next class.**
:::

## Intro

What is the life expectancy on the Island? Is it different for men and women?

We can't ask every islander, so we'll do what a previous class did: take a **stratified cluster sample** of the Island's records. Today we collect the data, and next class we analyze it.

Instead of computing life expectancy properly with a life table, we'll use the **mean age at death among islanders who have died** as a stand-in.

Here is the sampling plan:

1.  We stratify by **island**, so every island is represented.
2.  Within an island, the **villages** are the clusters. It would take too long to sample every village, so each of you is randomly assigned one.
3.  In your village, open the Town Hall, click **Records**, and open the **375-399** link. This lists every islander in your village who died between Island years 375 and 399.
4.  Randomly select **10** of those islanders and record their information in the shared spreadsheet.

::: callout-important
Your data goes into a spreadsheet that everyone in the class can see and edit. **Only add rows, never delete or overwrite anyone else's.** Enter `sex` as exactly `Female` or `Male`, and enter every value the same way each time.
:::

## Exercises

### Getting oriented

::: exercise
Log into [The Islands](https://islands.smp.uq.edu.au/index.php) with your College of Idaho email. Find the village you were assigned, and record its **name** and **island** here.

Before you start collecting: what **type of sampling** is this (simple random, stratified, cluster, or a combination)? Which part of the plan is the **stratification**, and which part is the **cluster**?
:::

::: answer
:::

### Collecting your sample

::: exercise
Open the **Records** for your village and click the **375-399** link. Then open [the shared Excel sheet](https://yotescollegeofidaho-my.sharepoint.com/:x:/g/personal/efriedlander_collegeofidaho_edu/IQB2Q0ijZYPzQYuumjlk_x09AQaTTMpIIkDa73TeNCMKJiI?e=16bSqH), create a new worksheet (the + button at the bottom), rename it with your island number, and copy and paste the list of islanders into it.

How many islanders are in your village's list? (Look at the number of rows.)
:::

::: answer
:::

::: exercise
Randomly select **10** islanders from your list by running the code below. Fill in your birthday in `MMDDYYYY` format (for example, `12181988`) and the number of islanders in your list.

The numbers it prints are the **row numbers** of the islanders you will sample.

```{r}
#| eval: false

birthday <- # put your birthday here in MMDDYYYY format (e.g. 12181988)
set.seed(birthday)
number_of_villagers <- # put your total number of islanders here
sample(1:number_of_villagers, 10)
```

Why do we use `set.seed()` here? Is choosing islanders this way a *random* sample?
:::

::: answer
```{r draw-sample}
#| eval: false

```
:::

::: exercise
Use the row numbers to look up your 10 islanders. In the **MASTER DATASET** worksheet of the same shared Excel sheet, add one row per islander with:

- your name
- the island
- the village
- the islander's name
- sex
- birth year
- death year

Remember: only add rows, and enter `sex` as exactly `Female` or `Male`.
:::

::: answer
:::

### Thinking about who is in the sample

::: exercise
We are only sampling islanders who died between Island years 375 and 399.

- Who **cannot** possibly appear in this sample?
- Why might this make our data a poor stand-in for the life expectancy of a *newborn* today?
- Name one way our sampling plan is better than just picking the first 10 islanders we come across.
:::

::: answer
:::

::: callout-important
Make sure all 10 of your islanders are entered in the MASTER DATASET by the start of the next class. Next time, we'll load the pooled data into R and analyze it.
:::