---
title: "AE 06: Comparing Models with the Coffee Truck"
subtitle: "Residuals, least squares, and R-squared"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

::: callout-important
- Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-06"
- Go to the assignment for this activity in [Canvas](<>) to get started.
- Upload the `ae-06-coffee-truck-eval.qmd` file into the folder you just created. We will generate three data sets for this activity shortly.
:::

## Recap

We're back at [Jo's coffee truck](https://www.stat2games.sites.grinnell.edu/games/coffeetruck.html). Today we'll fit and compare several models, and use $R^2$ to ask *how much* of the variation in `Sales` each one explains, and why that answer depends on more than just the model.

::: callout-important
This game's data lands in a table that's **public to anyone**, attached to your group name. Use the same team/Player Name from AE 05, or pick a new one that doesn't identify you.
:::

## Exercises

### Collecting three data sets

Every game keeps **Location = Business District**, **Time = Lunch**, **Weather = Relatively Constant**, and **Design Type = Random Sample** unless told otherwise below. On the ID screen, enter **`MATH2025FA26`** as the Group Name.

::: {.exercise}
Play a sim called **`music`**:

- **Music**: select all three levels (No Music, Hip Hop, Alternative).
- **Price**: fix at **\$3** (only one level selected).
- **Number of Repetitions**: `5`.
- Click **Start Sim**, then download the CSV. Upload it to your data folder as `music.csv`.
:::

::: {.answer}

:::

::: {.exercise}
Play a sim called **`price`**:

- **Price**: select all four levels (\$2, \$3, \$4, \$5).
- **Music**: fix at **No Music** (only one level selected).
- **Number of Repetitions**: `5`.
- Click **Start Sim**, then download the CSV. Upload it to your data folder as `price.csv`.
:::

::: {.answer}

:::

::: {.exercise}
Play a sim called **`both`**:

- **Music**: select all three levels.
- **Price**: select all four levels.
- **Number of Repetitions**: `2`.
- Click **Start Sim**, then download the CSV. Upload it to your data folder as `both.csv`.

Before moving on, check with your group: does your total number of days across all three sims still leave Jo with money in the bank? (`CurrentMoney` in the last row of each file.)
:::

::: {.answer}

:::

### Loading and cleaning

::: {.exercise}
Load the packages we'll use today, then read in all three data sets.
:::

::: {.answer}

```{r load-packages}
#| message: false
library(tidyverse)
library(mosaic)
library(ggformula)
library(broom)
library(coursekata)
library(yardstick)
```

```{r load-data}
#| message: false

```

:::

::: {.exercise}
Look at `price` (or `both`). You should notice `Price` was read in as a `chr` (character) column, not `dbl` (double).

- Why did that happen? Be specific.
- Fix it in both `price` and `both`, the same way we fixed `spend_per_resident_data` in HW 02: strip the `$` with `str_replace()`, then convert with `as.numeric()`. Call the new column `Price` (overwrite it).
:::

::: {.answer}

```{r clean-price}

```

:::

### Model 1: Price as a quantitative predictor

::: {.exercise}
Using `price`, plot `Sales` against `Price` with a fitted line (`gf_lm()`). Then fit `price_fit <- lm(Sales ~ Price, data = price)` and display it with `tidy()`.
:::

::: {.answer}

```{r plot-price}

```

```{r fit-price}

```

:::

::: {.exercise}
Use `augment()` to add residuals and fitted values to `price` for `price_fit`. Plot the residuals against the fitted values.

- Does the sum of the residuals look close to zero, like the least squares properties promised?
:::

::: {.answer}

```{r augment-price}

```

:::

::: {.exercise}
Compute `sd()` and `cor()` for `Price` and `Sales` by hand, and use $\hat\beta_1 = r\,s_Y/s_X$ and $\hat\beta_0 = \bar{y} - \hat\beta_1\bar{x}$ to reconstruct the slope and intercept. Confirm they match `price_fit`'s coefficients from `tidy()`.
:::

::: {.answer}

```{r hand-calc}

```

:::

### $R^2$, ANOVA, and RMSE

::: {.exercise}
Run `anova(price_fit)`. Identify SSModel and SSE from the output, add them to get SST, and use that to compute $R^2 = SSModel / SST$ by hand.

Then confirm your answer using `rsq()` on the augmented data. Also report `rmse()`.
:::

::: {.answer}

```{r anova-rsq}

```

:::

### Comparing across data sets

::: {.exercise}
Fit `lm(Sales ~ Music)` on `music`, and both `lm(Sales ~ Music)` and `lm(Sales ~ Price)` on `both`. For each of these three models, compute $R^2$ with `rsq()`.

Fill in the table below on the whiteboard (or here):

| Model | Data set | $R^2$ |
|:--|:--:|:--:|
| `Sales ~ Music` | `music` | |
| `Sales ~ Music` | `both` | |
| `Sales ~ Price` | `both` | |
| `Sales ~ Price` | `price` | |
:::

::: {.answer}

```{r compare-models}

```

:::

::: {.exercise}
Music's actual effect on `Sales` didn't change between `music` and `both` — you didn't change how the game responds to music, only what else you let vary. So why is $R^2$ for `Sales ~ Music` so different between those two data sets?

Hint: think about what happens to **SST** when price is also bouncing around.
:::

::: {.answer}

:::

::: {.exercise}
Based on this table, is it fair to say "the `Sales ~ Music` model is better than the `Sales ~ Price` model" without saying which data set each $R^2$ came from? Why or why not?
:::

::: {.answer}

:::

::: {.exercise}
*(If time permits)* Jo actually wants to maximize **profit**, not `Sales` — every data set the game gives you also has a `Profit` column. Does the model with the highest $R^2$ for `Sales` necessarily tell Jo the most about maximizing `Profit`? Explain.
:::

::: {.answer}

:::

::: callout-important
To submit the AE:

- Render the document to produce the HTML with all of your work from today's class.
- The driver for your group should upload your .qmd and .html files to the Canvas assignment.
:::
