---
title: "AE 10: Does Sex Explain More Variation Than Chance?"
subtitle: "Exploring the pooled Islands data and shuffling to build a null distribution"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

<!-- TODO (instructor): fill in the Canvas link and the URL of the cleaned CSV in the first exercise. -->

::: callout-important
- Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-10".
- Go to the assignment for this activity in [Canvas]() to get started.
- Upload the `ae-10-islands-inference.qmd` file into the folder you just created.
:::

## Intro

In AE 09 we collected a stratified cluster sample of the Islands. Today we work with the **pooled, already-cleaned** class data. (You'll do the cleaning yourself in **HW 05**.)

What is the life expectancy on the Island? Is it different for men and women? We'll use the **mean age at death among islanders who have died** as a stand-in for life expectancy, and then ask whether sex explains **more** variation in age at death than we'd expect by chance.

## Part 1: Exploring the pooled data

### Loading the data

::: exercise
Load the packages and the cleaned class data below. Use `glimpse()` to look at it. How many islanders are in the pooled sample? What does each row represent, and what variables do we have?
:::

::: answer
```{r load-data}
#| message: false
#| eval: false 
library(tidyverse)
library(mosaic)
library(ggformula)
library(broom)
library(coursekata)

# INSTRUCTOR: replace with the URL of the cleaned CSV
islands <- read_csv("CLEANED_ISLANDS_CSV_URL")
```
:::

### Estimating life expectancy

::: exercise
Explore `age_at_death`.

a.  Use `favstats()` to compute summary statistics for `age_at_death` overall, and then separately for each level of `sex`.

b.  Make a plot exploring the relationship between `sex` and `age_at_death`.

c.  In a sentence or two, report the **overall mean age at death**. This is our estimate of life expectancy on the Island. Then describe what the plot and the group summaries say about differences between men and women.
:::

::: answer
```{r eda}

```
:::

### Thinking critically

::: exercise
Let's think about what this sample can tell us.

- What **population** does our sample actually represent? Is it the whole Island? All islanders who have ever lived? Something else?
- Name **two ways** our estimate of life expectancy could be biased. (Think about who is *in* our sample, and who is missing.) One hint: think about who cannot possibly appear among "islanders who have died."
- Was this an **observational study** or an **experiment**? What could we conclude from comparing men and women in our sample, and what *couldn't* we conclude?
- Could you design an *experiment* on the Islands to find out whether sex affects life expectancy? Would you want to?
:::

::: answer
:::

::: callout-note
**Stop here.** We'll come back together and talk about what our results say (and don't say) before you continue with Part 2.
:::

## Part 2: Does sex explain more variation than chance?

::: exercise
How much variation in age at death does `sex` explain in our data?

a.  Compute the PRE (also called $R^2$) for the model predicting `age_at_death` from `sex` with `pre(age_at_death ~ sex, data = islands)`, and save it as `pre_obs`.

b.  Write a sentence interpreting `pre_obs` as a percent of the variation in age at death.

c.  Adding a predictor can never make PRE negative. Why not? Given that, why isn't "PRE is bigger than 0" convincing evidence that sex matters?
:::

::: answer
```{r pre-obs}

```
:::

::: exercise
Now let's imagine a world with **no relationship** between sex and age at death, just like the cards in class.

a.  Run `pre(shuffle(age_at_death) ~ sex, data = islands)` at least four times. Why does the answer change each time? Was it ever negative? Was it ever exactly 0?

b.  In your own words, what does `shuffle()` do to the relationship between `sex` and `age_at_death`? What does it *not* change?
:::

::: answer
```{r shuffle-once}

```
:::

::: exercise
Build the **null distribution** by shuffling 1000 times.

a.  Run `null_pre <- do(1000) * pre(shuffle(age_at_death) ~ sex, data = islands)`. Then make a histogram of the `pre` column of `null_pre` and add a vertical line at `pre_obs` (try `gf_histogram()` and `gf_vline()`).

b.  What does a single value in this histogram represent? What is the histogram a picture of?

c.  Compute the proportion of shuffles with a PRE at least as large as `pre_obs`: `mean(null_pre$pre >= pre_obs)`.

d.  Answer the question: does sex explain **more** variation in age at death than we'd expect if there were no relationship? Explain your reasoning, and say what your conclusion does and does *not* tell us (think back to how our sample was collected).
:::

::: answer
```{r null-dist}

```
:::

::: callout-important
To submit the AE:

- Render the document to produce the HTML with all of your work from today's class.
- Upload your .qmd and .html files to the Canvas assignment.
:::