---
title: "AE 11: What Makes a Song Danceable?"
subtitle: "Hypothesis tests and confidence intervals for the slope"
editor: visual
format:
  html:
    embed-resources: true
    include-in-header:
      text: |
        <style>
        /* Exercise / answer boxes. Canonical source: style/exercise-boxes.css
           in the course repo — copy from there if this ever drifts. */
        body { counter-reset: exercise; }
        .exercise, .answer {
          border-left: 4px solid #533860;
          padding: .6em 1em;
          margin: 1em 0;
          border-radius: 4px;
        }
        .exercise { background: #f2eef5; }
        .exercise::before {
          counter-increment: exercise;
          content: "Exercise " counter(exercise);
          display: block;
          font-weight: 700;
          color: #533860;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .answer {
          background: transparent;
          border-left-color: #a49363;
          border-left-style: dashed;
          min-height: 3.5em;
        }
        .answer::before {
          content: "Your answer";
          display: block;
          font-weight: 700;
          color: #8a7a4f;
          text-transform: uppercase;
          letter-spacing: .04em;
          font-size: .85em;
          margin-bottom: .35em;
        }
        .quarto-dark .exercise {
          background: rgba(255, 255, 255, .08);
          border-left-color: #FFCB70;
        }
        .quarto-dark .exercise::before {
          color: #FFCB70;
        }
        .quarto-dark .answer::before {
          color: #ffd283;
        }
        </style>
---

::: callout-important
- Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-11".
- Go to the assignment for this activity in [Canvas](https://cofi.instructure.com/courses/18743/assignments/244331) to get started.
- Upload the `ae-11-spotify-inference.qmd` and `spotify-popular.csv` files into the folder you just created.
:::

## Intro

What makes a song danceable? We'll analyze data on some of the most popular songs on Spotify (those with `track_popularity >= 80`) and ask whether a song characteristic helps predict its `danceability`.

Today we practice the full inference workflow for a slope:

1.  Choose a **test statistic** that captures the research question.
2.  Build its **null distribution** by shuffling, and notice that it looks normal.
3.  Standardize to get a $t$ statistic and a p-value from math instead of shuffling.
4.  Drop the "no relationship" assumption and use the **bootstrap** to build a **confidence interval**.

## Data

The data are a subset of the [Spotify Songs](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-21/readme.md) Tidy Tuesday data set, originally obtained from Spotify using the **spotifyr** R package. We'll focus on these variables:

| variable | class | description |
|:-----------------|:-----------------|:-----------------------------------|
| track_name | character | Song name |
| track_artist | character | Song artist |
| track_popularity | double | Song popularity (0-100), higher is better |
| danceability | double | How suitable a track is for dancing, from 0.0 (least danceable) to 1.0 (most danceable) |
| energy | double | Perceptual measure of intensity and activity, from 0.0 to 1.0. Fast, loud, noisy tracks have high energy |
| loudness | double | Overall loudness of a track in decibels (dB), typically between -60 and 0 |
| valence | double | Musical positiveness, from 0.0 (sad, angry) to 1.0 (happy, euphoric) |
| tempo | double | Estimated tempo in beats per minute (BPM) |
| acousticness | double | Confidence from 0.0 to 1.0 that the track is acoustic |
| liveness | double | Likelihood from 0.0 to 1.0 that the track was performed live (audience present) |
| duration_ms | double | Duration of the song in milliseconds |

```{r load-packages}
#| message: false
library(tidyverse)
library(ggformula)
library(broom)
library(knitr)
library(coursekata)  # for shuffle(), resample(), do(), b1()
library(openintro)   # for normTail()
```

```{r load-data}
#| message: false
spotify <- read_csv(here::here("data/spotify-popular.csv"))
```

### Predictor assignment

Each group is assigned a different explanatory variable. Our response variable is always `danceability`.

- **Group 1:** `energy`
- **Group 2:** `loudness`
- **Group 3:** `valence`
- **Group 4:** `tempo`
- **Group 5:** `acousticness`
- **Group 6:** `liveness`

## Part 1: Exploring and fitting

::: exercise
Make a visualization of your explanatory variable, a visualization of `danceability`, and a scatterplot of `danceability` against your explanatory variable. Describe the relationship you see.
:::

::: answer
```{r eda}
```
:::

::: exercise
Fit a linear model predicting `danceability` from your explanatory variable. Save it as `spotify_fit` and use `tidy()` and `kable()` to display it. Write out the fitted equation and interpret the slope in context. Have your reporter write your equation on the whiteboard.
:::

::: answer
```{r fit-model}

```
:::

## Part 2: Hypothesis testing

::: exercise
Write your **research question** in words. Which **test statistic** captures this question? Then write the null and alternative hypotheses in words and in symbols. Finally, list every factor you can think of that will affect how convincing your observed statistic is (without computing anything).
:::

::: answer
:::

::: exercise
Using math instead of shuffling.

a.  Compute the test statistic $T = (b_1 - 0)/SE_{b_1}$ using the output of `tidy()`. Interpret it: how many standard errors is your slope from the null value, and in which direction?

b.  In the output, find the p-value and then sketch the $t$ distribution and shade your p-value.

c.  How strong is the evidence against $H_0$? Write a conclusion in the context of the problem. How does it compare to your simulation p-value from the previous exercise?
:::

::: answer
```{r t-stat}

```
:::

::: callout-important
To submit the AE:

- Render the document to produce the HTML with all of your work from today's class.
- Upload your .qmd and .html files to the Canvas assignment.
:::